---
title: "Gene names: old identifiers and Excel errors"
description: >
  Convert original Ensembl IDs from a 2014 single-cell study and repair
  Excel-damaged gene symbols with annotation-specific mappings.
output:
  rmarkdown::html_vignette:
    self_contained: true
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Gene names: old identifiers and Excel errors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(seqout)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  eval = requireNamespace("genevintage", quietly = TRUE)
)
example_dir <- system.file("extdata", "gene-names", package = "seqout")
read_example <- function(name, ...) {
  read.delim(file.path(example_dir, name), check.names = FALSE, ...)
}
```

`SeqoutGeneNames()` converts gene IDs to names using **genevintage**.
`SeqoutCorrectNames()` repairs gene symbols that Excel changed into dates.
Both accept the result of `SeqoutMatrix()` or a numeric matrix with genes on rows.
They preserve expression values and row order, keep feature metadata aligned,
and record the original labels and conversion status.

Install seqout and its genevintage dependency from Saket Lab's r-universe:

```{r install, eval=FALSE}
install.packages("seqout", repos = c(
  "https://saketlab.r-universe.dev", "https://cloud.r-project.org"
))
```

The full-data commands below download GEO files and annotation tables on first
use. The evaluated examples use small, unaltered excerpts bundled with seqout,
so they also run offline. Their source URLs and extraction details are in
`extdata/gene-names/README.md` inside the installed package.

## Old Ensembl IDs from a 2014 single-cell experiment

[GSE52529](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE52529), the
Monocle myoblast differentiation study, became public on 23 March 2014.
Its single-cell file has 47,192 gene rows and 372 cells, with the original
versioned Ensembl IDs such as `ENSG00000000003.10`. The values are FPKM, not raw
counts. Select `GSE52529_fpkm_matrix.txt.gz`; the separate
`truseq` file contains bulk measurements.

```{r old-study-download, eval=FALSE}
old <- SeqoutListCounts("GSE52529") |>
  SeqoutMatrix(sample = "GSE52529_fpkm_matrix.txt.gz")
dim(old$X)
head(rownames(old$X))

# automatic selection; decision recorded below
automatic <- old |> SeqoutGeneNames()
automatic$gene_annotation[[1]]$annotation
```

Automatic release selection is a ranked estimate: filtered gene sets can fit
several references, adjacent releases can share identical gene sets, and
versioned IDs do not uniquely identify the annotation used. For this file the
detector selects GENCODE 17 on GRCh37, matching the
[sample's processing record](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM1268960).
You do not need to supply the release or assembly. All 47,192 deposited IDs,
including their version suffixes, match that reference exactly.

```{r automatic-audit, eval=FALSE}
head(SeqoutGeneReport(automatic))
stopifnot(
  identical(unname(old$X), unname(automatic$X)),
  identical(rownames(automatic$X), rownames(automatic$var))
)
```

Ranking releases needs no annotation download. It uses the complete original ID
vector, bundled separately from the expression excerpt, because ranking only the
displayed genes would change the evidence.

```{r old-study-ranking}
id_file <- gzfile(file.path(example_dir, "GSE52529-gene-ids.txt.gz"))
old_ids <- readLines(id_file)
close(id_file)
head(genevintage::detect_release(old_ids), 3)
```

Update genevintage if your installation still ranks release 25 first for this
file.

The first 12 genes and first three cells of the deposited file, converted with
the matching GENCODE 17 excerpt:

```{r old-study-offline}
old_excerpt <- as.matrix(read_example("GSE52529-excerpt.tsv.gz", row.names = 1))
gencode17 <- read_example("gencode17-excerpt.tsv.gz")
old_named <- SeqoutGeneNames(old_excerpt, mapping = gencode17)
SeqoutGeneReport(old_named)[, c("input", "name", "status")]
stopifnot(identical(as.vector(old_excerpt), as.vector(old_named)))
```

Version suffixes are ignored for matching but retained in the audit. For example,
`ENSG00000000003.10` becomes `TSPAN6`. An ID without a name in the selected
annotation stays as its input ID. Duplicate symbols receive distinct row labels;
the audit's `name` column preserves the unsuffixed symbol. No rows are summed.

## Correct Excel-damaged names in a real deposited matrix

[GSE101942](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE101942) has
Excel-damaged gene names. We verified that the
original `GSE101942_IPSC_rawCounts.txt.gz` has 56,637 rows (five of them
counting-summary rows) and six samples, with `1-Mar`, `9-Sep`, and `15-Sep`
among the row labels.

Select the raw-count file explicitly; the study also deposits differential
expression result tables, which are not count matrices.

```{r excel-study-download, eval=FALSE}
damaged <- SeqoutListCounts("GSE101942") |>
  SeqoutMatrix(sample = "GSE101942_IPSC_rawCounts.txt.gz")

# explicit target, not automatic
repaired <- damaged |> SeqoutCorrectNames(species = "human", release = 116)
report <- SeqoutGeneReport(repaired)
report[report$status != "unchanged", c("input", "name", "status")]
stopifnot(identical(unname(damaged$X), unname(repaired$X)))
```

Against Ensembl 116, 25 labels are corrected. `15-Sep` stays unresolved because
this annotation has no matching candidate for it.

The bundled excerpt contains all 26 damaged labels plus `TP53` as an unchanged
control, with the original six sample columns and their counts. The
annotation excerpt spans the reference's complete month-related candidate
vocabulary.

```{r excel-study-offline}
excel_excerpt <- as.matrix(read_example("GSE101942-excerpt.tsv.gz", row.names = 1))
human116 <- read_example("ensembl116-excel.tsv.gz")
excel_fixed <- SeqoutCorrectNames(excel_excerpt, mapping = human116, quiet = TRUE)
excel_report <- SeqoutGeneReport(excel_fixed)
excel_report[
  excel_report$input %in% c("1-Mar", "9-Sep", "15-Sep", "TP53"),
  c("input", "name", "status")
]
table(excel_report$status)
stopifnot(identical(as.vector(excel_excerpt), as.vector(excel_fixed)))
```

### Historical versus newer symbols

The same damaged labels resolve differently against human Ensembl 95 and 116.
This compares two reference vocabularies; it does not infer which release
GSE101942 used.

```{r excel-vintages}
human95 <- read_example("ensembl95-excel.tsv.gz")
excel_old <- SeqoutCorrectNames(excel_excerpt, mapping = human95, quiet = TRUE)
old_report <- SeqoutGeneReport(excel_old)
show <- excel_report$input %in% c("1-Mar", "9-Sep")
data.frame(
  deposited = excel_report$input[show],
  ensembl95 = old_report$name[show],
  ensembl116 = excel_report$name[show]
)
```

`1-Mar` resolves to `MARCH1` in Ensembl 95 and `MARCHF1` in 116; `9-Sep` to
`SEPT9` and `SEPTIN9`. Names alone cannot tell seqout which species or vintage to
choose, so Excel correction requires either an annotation table or both
`species` and `release`. Ambiguous dates are retained and marked `ambiguous`;
tokens without candidates are marked `unresolved`. genevintage cannot repair
some formats or losses of numeric precision. Review the report before analysis,
and correct only columns known to hold gene labels.

If the damaged symbols are in feature metadata, pass that column explicitly:

```{r feature-column, eval=FALSE}
repaired <- SeqoutCorrectNames(
  rna,
  ids = rna$var$symbol, species = "human", release = 116
)
```

The original `var$symbol` column is retained. The repaired labels become the row
names of both `X` and `var`, and each call appends its audit to
`gene_annotation`. Save the corrected `seqout_matrix` if you need to retain
this audit alongside a later Seurat or SingleCellExperiment conversion.
