---
title: "Get Started with seqout"
description: >
  Learn the core workflows of the seqout package: search genomic databases, 
  retrieve study metadata, read counts matrices, download files, and select backends.
output:
  rmarkdown::html_vignette:
    self_contained: true
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Getting started with seqout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The `seqout` R package provides a client interface for [seqout.org](https://seqout.org). Use this package to query and retrieve study-level and sample-level metadata from seven major public genomic archives:
*   Gene Expression Omnibus (GEO)
*   Sequence Read Archive (SRA)
*   European Nucleotide Archive (ENA)
*   ArrayExpress
*   DNA Data Bank of Japan (DDBJ DRA and GEA)
*   Genome Sequence Archive (GSA)

This guide introduces the primary features of the package. Each section links to a dedicated article for in-depth information.

## Installation

Install the package using the `pak` manager:

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("saketlab/seqout/R")
```

By default, the package connects to the REST API backend. This backend is always up-to-date and does not require an API key or local setup.

```{r setup}
library(seqout)
```

## Search the archives

The `seqout` package provides search functions to query either study-level or sample-level metadata.

### Query studies

To search across all seven supported genomic archives simultaneously, use `SeqoutSearch()`. You can supply free-text queries, specific filters, or both.

```{r search}
SeqoutSearch("cervical cancer single cell", db = "geo", limit = 5)
```

### Query samples (cohorts)

To search for individual samples rather than entire studies, use `SampleSearch()`. This function queries the harmonized metadata generated by `seqout.org`. Because free-text attribute fields are standardized into a unified vocabulary, you can filter by tissue, disease, or age consistently across different studies.

```{r cohort}
liver <- SampleSearch(
  organism = "Homo sapiens", 
  tissue = "liver", 
  age_min_years = 50,
  limit = 5
)
liver[, c("sample", "study_accession", "tissue", "disease", "sex", "age")]
```

For more details on query operators and advanced filters, see [Search](articles/search.html).

## Retrieve study metadata

Use `SeqoutGet()` to fetch detailed metadata for a specific study, sample, or run accession. The function automatically maps the query accession to related records in other archives. 

To save network bandwidth, the returned dataset object uses lazy loading. It queries the API and caches the results only when you access a specific field for the first time.

```{r dataset}
d <- SeqoutGet("GSE168652")
d
```

Access specific metadata fields directly:

```{r dataset-fields}
d$meta$title
c(samples = nrow(d$samples), runs = nrow(d$runs), papers = nrow(d$pubs))
```

### Harmonized sample attributes

The `enriched` field contains the harmonized cell-level and sample-level metadata (such as standardized tissue types, disease status, and assay methods) generated for each sample:

```{r enriched}
d$enriched[, c("sample", "tissue", "disease", "assay")]
```

For more information on the available metadata fields, see [Metadata](articles/metadata.html).

## Read counts matrices

You can resolve, download, and parse processed supplementary files directly into R matrices.

### List counts files

Use `SeqoutCounts()` to retrieve a table of supplementary files for a GEO accession. The function groups related files into **units**, which represent the smallest collection of files that `seqout` can parse as a single matrix (e.g., a 10x Genomics triplet, a `.h5` file, a `.h5ad` file, or a `.rds` file).

This function queries metadata only and does not download the actual matrix files, which can exceed tens of gigabytes in size.

```{r counts-table}
counts <- SeqoutCounts("GSE297547")
m <- counts

nrow(m)
head(m[, c("unit", "sample", "format", "preferred", "has_metadata")], 4)
```

### Load a matrix

To download and parse a specific matrix unit into R, use `SeqoutMatrix()`. The returned list contains the expression matrix in `X`, cell annotations in `obs`, and feature annotations in `var`.

```{r counts-read, eval = FALSE}
rna <- SeqoutMatrix(counts, sample = "GSM8994520")
dim(rna$X)
```

### Convert to Seurat

To construct a Seurat object directly from a matrix unit or a GSM accession, use `Seqout2Seurat()`. Additional arguments are forwarded to `CreateSeuratObject()`.

```{r counts-seurat, eval = FALSE}
obj <- Seqout2Seurat("GSM8994520", min.cells = 3, min.features = 200)
```

For preferred unit selection rules and Seurat conversion workflows, see [Counts](articles/counts.html).

## Download data files

The package provides three functions to download raw and processed files to disk:
*   `DownloadSupplementary()`: Downloads processed dataset files (such as counts matrices, annotations, and compressed archives).
*   `DownloadRuns()`: Downloads raw sequencing reads (FASTQ or SRA format).
*   `DownloadBams()`: Downloads submitted alignment files (BAM format).

By default, these functions save files in a subdirectory named after the target accession. They automatically verify checksums and skip files that are already present on disk.

To inspect the files associated with a dataset before downloading them, check the `supplementary` field:

```{r supplementary}
head(d$supplementary, 4)
```

To download the files, call the respective functions:

```{r downloads, eval = FALSE}
DownloadSupplementary("GSE168652")
DownloadRuns("SRR12012336")
DownloadBams("ERP117016")
```

For more details on file download modes and checksum verification, see [Downloads](articles/downloads.html).

## Choose a backend

The package supports two data retrieval backends:

| Backend | Mechanism | Best Use Cases |
| --- | --- | --- |
| `"api"` (Default) | Queries the `seqout.org` REST API over HTTP. | Standard workflows. The data is always up-to-date. |
| `"parquet"` | Queries the published Parquet database dump using DuckDB. | Offline workflows, large batch queries, and custom SQL analytics. |

To use the Parquet backend, you must configure the connection explicitly. This requirement prevents accidental, high-bandwidth scans of multi-gigabyte files.

```{r parquet}
con <- SeqoutConnect("parquet", quiet = TRUE)

Query("SELECT count(*) AS n FROM geo_series", con = con)
```

You can pass the connection object to functions using the `con` argument, or make it the default connection for your session by calling `SeqoutDefault(con)`.

To close the Parquet connection and release system resources, call `SeqoutClose()`:

```{r close}
SeqoutClose(con)
```

For instructions on downloading Parquet data dumps for local, high-performance offline queries, see [Downloads](articles/downloads.html).

## Next steps

Explore the following guides to learn more about `seqout`:
*   [Search](articles/search.html) — Query operators, filters, and sample cohort construction.
*   [Metadata](articles/metadata.html) — Dataset structures, publication records, and read-derived metrics.
*   [Counts](articles/counts.html) — GEO supplementary file parsing and Seurat object construction.
*   [Downloads](articles/downloads.html) — Downloading raw reads, alignments, and database dumps.
