---
title: "Get Started with seqout"
description: >
  Core seqout workflows: search genomic databases, retrieve study metadata, read counts matrices, download files, and select backends.
output:
  rmarkdown::html_vignette:
    self_contained: true
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Getting started with seqout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`seqout` is an R client for [seqout.org](https://seqout.org). It queries study-level and sample-level metadata from public genomic archives:
*   Gene Expression Omnibus (GEO)
*   Sequence Read Archive (SRA)
*   European Nucleotide Archive (ENA)
*   ArrayExpress
*   DNA Data Bank of Japan (DDBJ DRA and GEA)
*   Genome Sequence Archive (GSA)

This guide covers the main workflows and links to task-specific articles.

## Installation

Install the package using the `pak` manager:

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("saketlab/seqout/R")
```

By default, the package connects to the REST API. It needs no API key or local setup.

```{r setup}
library(seqout)
```

## Search the archives

`seqout` can search study-level or sample-level metadata.

### Query studies

Use `SeqoutSearch()` to search all archives. Supply free-text queries, filters, or both.

```{r search}
SeqoutSearch("cervical cancer single cell", db = "geo", limit = 5)
```

### Query samples (cohorts)

Use `SampleSearch()` to search individual samples. Harmonized metadata supports filters such as tissue, disease, and age across studies.

```{r cohort}
liver <- SampleSearch(
  organism = "Homo sapiens",
  tissue = "liver",
  age_min_years = 50,
  limit = 5
)
liver[, c("sample", "study_accession", "tissue", "disease", "sex", "age")]
```

For query operators and filters, see [Search](articles/search.html).

### Everything pipes

Each function takes its subject first, so calls chain with R's native pipe: an
accession, a `seqout_counts` table, a list of cohort filters, or the connection
itself when the call has no other subject.

```{r pipes}
# A PubMed ID or a DOI, told apart by the value, to the datasets that cite it
"34764296" |>
  Paper() |>
  nrow()

list(tissue = "liver", organism = "Homo sapiens") |>
  SampleSearch(limit = 3) |>
  subset(select = c(sample, study_accession, disease))
```

`con` is the last argument everywhere else.

## Retrieve study metadata

Use `SeqoutGet()` to fetch metadata for a study, sample, or run accession. The function maps the accession to related records in other archives. 

The returned dataset uses lazy loading: it queries the API and caches results when you read a field.

```{r dataset}
d <- SeqoutGet("GSE168652")
d
```

Access specific metadata fields directly:

```{r dataset-fields}
d$meta$title
c(samples = nrow(d$samples), runs = nrow(d$runs), papers = nrow(d$pubs))
```

### Harmonized sample attributes

The `enriched` field contains harmonized cell-level and sample-level metadata, such as tissue, disease, and assay labels:

```{r enriched}
d$enriched[, c("sample", "tissue", "disease", "assay")]
```

For metadata fields, see [Metadata](articles/metadata.html).

## Read counts matrices

`seqout` can resolve, download, and parse processed GEO supplementary files into R matrices.

### List counts files

Use `SeqoutListCounts()` to list matrix units for a GEO accession. A unit is the smallest file group that `seqout` can parse as one matrix, such as a 10x triplet, `.h5`, `.h5ad`, or `.rds` file.

This function queries metadata and downloads no matrix files.

```{r counts-table}
counts <- SeqoutListCounts("GSE297547")
m <- counts

nrow(m)
head(m[, c("unit", "sample", "format", "preferred", "has_metadata")], 4)
```

### Load a matrix

Use `SeqoutMatrix()` to read one matrix unit. The returned list contains `X`, `obs`, and `var`.

```{r counts-read, eval = FALSE}
rna <- SeqoutMatrix(counts, sample = "GSM8994520")
dim(rna$X)
```

### Convert to Seurat

Use `Seqout2Seurat()` with a matrix unit, GSM accession, GSE accession, or `SeqoutListCounts()` table. Extra arguments go to `CreateSeuratObject()`.

```{r counts-seurat, eval = FALSE}
obj <- Seqout2Seurat("GSE297547", sample = "GSM8994520", min.cells = 3, min.features = 200)
```

For unit selection and Seurat workflows, see [Counts](articles/counts.html).

## Download data files

Three functions download raw and processed files:
* `DownloadSupplementary()`: processed dataset files, such as counts matrices, annotations, and archives.
* `DownloadRuns()`: raw sequencing reads in FASTQ or SRA format.
* `DownloadBams()`: submitted BAM alignment files.

These functions save files in a directory named after the accession, verify checksums, and skip files already present on disk.

Check the `supplementary` field before downloading:

```{r supplementary}
head(d$supplementary, 4)
```

Download with:

```{r downloads, eval = FALSE}
DownloadSupplementary("GSE168652")
DownloadRuns("SRR12012336")
DownloadBams("ERP117016")
```

For download modes and checksums, see [Downloads](articles/downloads.html).

## Choose a backend

The package supports two backends:

| Backend | Mechanism | Best Use Cases |
| --- | --- | --- |
| `"api"` (default) | Queries the `seqout.org` REST API over HTTP. | Standard workflows. |
| `"parquet"` | Queries the published Parquet dump using DuckDB. | Offline workflows, large batch queries, and SQL analytics. |

Configure the Parquet backend explicitly to avoid accidental dump scans.

```{r parquet}
con <- SeqoutConnect("parquet", quiet = TRUE)

Query("SELECT count(*) AS n FROM geo_series", con = con)
```

Pass the connection through `con`, or make it the session default with `SeqoutDefault(con)`.

Close the Parquet connection with `SeqoutClose()`:

```{r close}
SeqoutClose(con)
```

For Parquet downloads, see [Downloads](articles/downloads.html).

## Next steps

Next guides:
* [Search](articles/search.html): query operators, filters, and sample cohort construction.
* [Metadata](articles/metadata.html): dataset structures, publication records, and read-derived metrics.
* [Counts](articles/counts.html): GEO supplementary file parsing and Seurat object construction.
* [Downloads](articles/downloads.html): raw reads, alignments, and database dumps.
