---
title: "Getting started with seqout"
description: >
  Get datasets from GEO, SRA, ENA, ArrayExpress, DDBJ and GSA.
output:
  rmarkdown::html_vignette:
    self_contained: true
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Getting started with seqout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
api_reachable <- tryCatch(
  curl::curl_fetch_memory(
    "https://seqout.org/api/project/GSE151530",
    curl::new_handle(timeout = 30)
  )$status_code == 200L,
  error = function(e) FALSE
)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = api_reachable
)
```

`seqout` is an R client for [seqout.org](https://seqout.org). It gets metadata
about high-throughput sequencing datasets from seven public repositories: GEO,
SRA, ArrayExpress, ENA, GSA, DRA and GEA.

Use it to search the repositories, to read a study and its samples, and to load
GEO supplementary files as counts matrices.

## Installation

```{r install, eval=FALSE}
# install.packages("pak")
pak::pak("saketlab/seqout/R")
```

## A quick example

Search the repositories:

```{r quick-search}
library(seqout)

seqout_search("lung cancer single cell", db = "geo", limit = 5)
```

Then open a dataset. `seqout_get()` takes any accession and finds the related
records across the archives: the metadata, the samples, the runs and the
papers.

```{r quick-dataset}
d <- seqout_get("GSE168652")
c(title = d$meta$title, samples = nrow(d$samples), runs = nrow(d$runs))
```

## Two backends

`seqout` can get data in two ways.

| Backend | How it works | When to use it |
| --- | --- | --- |
| REST API (default) | It calls the seqout.org web API. | Normal use. It is always up to date. |
| Parquet | It reads Parquet data dumps with DuckDB. | Offline or local use, SQL, and large batch jobs. |

The REST API is the default, and it needs no setup. The examples above show
this: they call no other function first.

Select the Parquet backend with `seqout_connect()`. It reads the Parquet files
directly, from a remote server or from a local directory. It needs the `duckdb`
package.

```{r connect}
con <- seqout_connect("parquet")
con
```

Give the connection to a function with `con =`. Or make it the default for the
session:

```{r default}
seqout_default(con)
```

The sections below read Parquet because of that call. `seqout_default(NULL)`
selects the REST API again.
## Search

`seqout_search()` is the only search function. Give it a query, filters, or both.

```{r search}
seqout_search("liver cancer scRNA", organism = "Homo sapiens", limit = 10)
```

For the filters, the sort order and the cost of a large result set, read
[Searching](searching.html).

## Get detailed metadata for a project

`GSE151530` is a single-cell study of liver cancer with many citations.
`project()` gives you the full record: the title, the description, the
contributors, the organisms, the instruments and the supplementary files.

```{r metadata-full}
meta <- project("GSE151530")
meta
```

A project record has 25 columns. That is too wide to read. `transpose = TRUE`
gives you one row for each field:

```{r metadata-transposed}
project("GSE151530", transpose = TRUE)
```

**Samples** in this project:

```{r samples}
samples <- project_samples("GSE151530")
samples
```


## Enriched (ontology-annotated) metadata

SeqOut adds ontology terms to the samples. The terms give the tissue, the
disease, the cell type and the assay. `project_enriched()` reads them:

```{r enriched}
enriched <- project_enriched("GSE151530")
enriched
```

The result has a column for each term, with the ontology ID and the confidence
score of that term:

```{r enriched-inspect}
library(dplyr)

enriched |>
  select(sample, tissue, disease, cell_type, assay, assay_category, sample_type) |>
  head(10)
```


## Publications

`publications()` reads the publication record of one or more accessions. The
record has the DOI, the PubMed ID, the journal and the citation count.

```{r publications}
publications(c("GSE151530", "GSE98638"))
```

`project_citations()` gives the same data for one project:

```{r citations-tibble}
project_citations("GSE151530")
```


## BibTeX export

Give `format = "bibtex"` to get a BibTeX string:

```{r bibtex}
bib <- project_citations("GSE151530", format = "bibtex")
cat(bib)
```

This produces entries like:

```
@article{Lichun2019,
  author  = {Lichun Ma, Maria O Hernandez, ...},
  title   = {Tumor Cell Biodiversity Drives Microenvironmental ...},
  journal = {Cancer cell},
  year    = {2019},
  doi     = {10.1016/j.ccell.2019.08.007},
  pmid    = {31588021}
}
```

To save to a `.bib` file:

```{r save-bibtex, eval=FALSE}
writeLines(bib, "GSE151530.bib")
```


## Using SQL and dplyr directly

All the above functions run on DuckDB under the hood. You can also query
the tables directly with SQL:

```{r sql}
query("
  SELECT dominant_scientific_name AS organism, count(*) AS n_projects
  FROM unified_metadata
  WHERE dominant_scientific_name IS NOT NULL
  GROUP BY organism
  ORDER BY n_projects DESC
  LIMIT 10
")
```

`dplyr` and `dbplyr` also work, and they read the data only when you ask for it.
`tbl()` reads the DuckDB catalog, so the view must exist first:

```{r dplyr}
library(dbplyr)

register_tables("unified_metadata")

tbl(con$db, "unified_metadata") |>
  filter(
    dominant_scientific_name == "Homo sapiens",
    country == "Japan"
  ) |>
  count(dominant_assay_l2, sort = TRUE) |>
  head(10) |>
  collect()
```


## Cache tables locally

`cache_table()` copies a remote view into local DuckDB storage. Use it when you
read the same table more than once:

```{r cache}
cache_table("unified_metadata")

# Subsequent queries hit local storage
query("SELECT count(*) AS n FROM unified_metadata_local")

# Clean up
clear_cache()
```


## Close connection

Only the Parquet backend holds anything worth closing. Put the REST default
back and release the DuckDB handle:

```{r close}
seqout_default(NULL)
seqout_close(con)
```
