---
title: "Getting started with seqout"
description: >
  A short tour of the package: search the archives, read the metadata, load a
  counts matrix, download files, and select a backend.
output:
  rmarkdown::html_vignette:
    self_contained: true
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Getting started with seqout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`seqout` is an R client for [seqout.org](https://seqout.org). It can query and 
read metadata of high-throughput sequencing studies from seven public archives: 
GEO, SRA, ENA, ArrayExpress, DDBJ (DRA and GEA) and GSA.

This page shows each part of the package in a few lines. Each section links to
a longer article.

```{r install, eval = FALSE}
# install.packages("pak")
pak::pak("saketlab/seqout/R")
```

The REST API is the default backend. It needs no setup and no key.

```{r setup}
library(seqout)
```

## Search

`SeqoutSearch()` searches the seven archives at the same time. Give it a
query, filters, or both. You can learn more about [how search works](https://seqout.org/howsearchworks).

```{r search}
SeqoutSearch("cervical cancer single cell", db = "geo", limit = 5)
```

`SampleSearch()` searches the samples instead of the studies. It reads the
harmonised data, which is what seqout writes after it reads each sample's free
text, so one filter can name a tissue, a disease or an age across every study.

```{r cohort}
liver <- SampleSearch(
  organism = "Homo sapiens", tissue = "liver", age_min_years = 50,
  limit = 5
)
liver[, c("sample", "study_accession", "tissue", "disease", "sex", "age")]
```

Read [Search](articles/search.html) for the full filter set, the sort order and
the cost of a large result set.

## Read the metadata

`SeqoutGet()` takes any accession. It finds the related records in the other
archives itself. Each field makes its request the first time you read it, and
it keeps the answer.

```{r dataset}
d <- SeqoutGet("GSE168652")
d
```

```{r dataset-fields}
d$meta$title
c(samples = nrow(d$samples), runs = nrow(d$runs), papers = nrow(d$pubs))
```

The `enriched` field holds the harmonised data for each sample.

```{r enriched}
d$enriched[, c("sample", "tissue", "disease", "assay")]
```

Read [Metadata](articles/metadata.html) for the other fields, the publication
records and the read-derived calls.

## From GEO to a counts matrix

`SeqoutCounts()` groups the supplementary files of a GEO accession into units.
A unit is a group of files that read as one matrix. This call downloads
nothing, because a GEO payload can be tens of gigabytes.

```{r counts-manifest}
counts <- SeqoutCounts("GSE297547")
m <- Manifest(counts)

nrow(m)
head(m[, c("unit", "sample", "format", "preferred", "has_metadata")], 4)
```

`SeqoutMatrix()` downloads one unit and reads it. The result holds the matrix
in `X`, the cell annotation in `obs` and the gene annotation in `var`.

```{r counts-read, eval = FALSE}
rna <- SeqoutMatrix(counts, sample = "GSM8994520")
dim(rna$X)
```

This vignette does not run that call, because the unit is 300 MB.

Read [Counts](articles/counts.html) for the unit rules, the choice of assay and
the path into Seurat.

## Download the files

The `supplementary` field lists the processed files of a study, one row per
file.

```{r supplementary}
head(d$supplementary, 4)
```

Three functions write files to disk. `DownloadSupplementary()` takes the
processed files, `DownloadRuns()` takes the reads, and `DownloadBams()` takes
the alignments the submitter sent.

```{r downloads, eval = FALSE}
DownloadSupplementary("GSE168652")
DownloadRuns("SRR12012336")
DownloadBams("ERP117016")
```

Read [Downloads](articles/downloads.html) for the file choice, the checksums
and the buckets that no anonymous client can read.

## The two backends

| Backend | How it works | When to use it |
| --- | --- | --- |
| `"api"` (the default) | It calls the seqout.org REST API. | Normal use. It is always current. |
| `"parquet"` | It reads the Parquet dump with DuckDB. | SQL, offline work and large batch jobs. |

You must select the Parquet backend. The package does not change to it for
you, because one lookup would become a scan of many gigabytes. It needs the
`duckdb` package.

```{r parquet}
con <- SeqoutConnect("parquet", quiet = TRUE)

Query("SELECT count(*) AS n FROM geo_series", con = con)
```

Give the connection to a function with `con =`, or make it the default for the
session with `SeqoutDefault(con)`.

By default the Parquet backend reads the published dump over HTTP.
`DownloadDump()` fetches those files, and `data_dir` points a connection at
them. That is what makes offline work and a large batch job fast.

```{r data-dir, eval = FALSE}
DownloadDump("~/seqout-dump", tables = c("unified_metadata", "geo_series"))

con <- SeqoutConnect("parquet", data_dir = "~/seqout-dump")
```

The whole dump is tens of gigabytes, so `DownloadDump()` reports the size
before it starts and takes a `tables` argument. A file already on disk is left
alone, so a second call finishes an interrupted first one.

```{r close}
SeqoutClose(con)
```

The two backends do not hold the same data. `SeqoutSearch()`,
`SampleSearch()` and the read-derived tables need the REST API. SQL and the
local cache need Parquet.

## Where to go next

* [Search](articles/search.html) — the filters, the sort order and the cohorts.
* [Metadata](articles/metadata.html) — the fields of a dataset and the papers.
* [Counts](articles/counts.html) — GEO supplementary files as a matrix.
* [Downloads](articles/downloads.html) — reads, alignments and processed files.
