---
title: "Getting started with seqout"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with seqout}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

## Installation

```{r install, eval=FALSE}
# install.packages("pak")
pak::pak("saketlab/seqout/R")
```

## Connect to SeqOut

All queries start by creating a connection. This sets up a local DuckDB
instance that reads remote Parquet files from `seqout.org`, a full download
is not required.

```{r connect}
library(seqout)

con <- seqout_connect()
con
```

## Search for datasets

Suppose we are interested in finding single-cell RNA-seq datasets
related to liver cancer. `find_projects()` lets us search by
keywords and apply structured filters — no SQL needed:

```{r search-liver}
results <- find_projects(con,
  keywords = "liver cancer scRNA",
  organism = "Homo sapiens",
  sort_by  = "citations",
  limit    = 10
)
results
```

Keywords are matched with AND logic: every word must appear somewhere in the
project title or description. We can further narrow the search by adding
filters:

```{r search-narrow}
# Only GEO projects from 2020 onwards with at least 10 samples
find_projects(con,
  keywords    = "liver cancer scRNA",
  organism    = "Homo sapiens",
  source      = "geo",
  year_from   = 2020,
  min_samples = 10,
  sort_by     = "citations"
)
```

### Discover valid filter values

Not sure what to pass for `assay` or `organism`? The `list_*` helpers
return valid values with project counts:

```{r list-values}
# Specific assay types (RNA-seq, ChIP-seq, WGS, ...)
list_assays(con)

# Broad assay categories (Transcriptomic, Genomic Sequencing, ...)
list_assays(con, level = "broad")

# Top organisms
list_organisms(con, limit = 10)

# Top countries
list_countries(con, limit = 10)
```


## Get detailed metadata for a project

From our search above, `GSE151530` is a highly cited single-cell liver
cancer study. Let's get its metadata.

**Title and description:**

```{r metadata-basic}
project_metadata(con, "GSE151530")
```

**Full metadata** including contributors, organisms, instruments, and
supplementary data links:

```{r metadata-full}
meta <- project(con, "GSE151530")
meta
```

**Samples** in this project:

```{r samples}
samples <- project_samples(con, "GSE151530")
samples
```


## Enriched (ontology-annotated) metadata

SeqOut annotates samples with standardized ontology terms for tissue,
disease, cell type, assay, and more. `project_enriched()` retrieves
these annotations:

```{r enriched}
enriched <- project_enriched(con, "GSE151530")
enriched
```

The enriched metadata includes columns like `tissue`, `disease`, `cell_type`,
`assay`, `assay_category`, and `sample_type`, along with their
corresponding ontology IDs and confidence scores:

```{r enriched-inspect}
library(dplyr)

enriched |>
  select(sample, tissue, disease, cell_type, assay, assay_category, sample_type) |>
  head(10)
```


## Publications

`publications()` parses the structured publication data — DOI, PubMed ID,
journal, and citation count — for one or more accessions:

```{r publications}
publications(con, c("GSE151530", "GSE98638"))
```

For a single project, `project_citations()` returns the same data:

```{r citations-tibble}
project_citations(con, "GSE151530")
```


## BibTeX export

To get a ready-to-use BibTeX string for a project's publications, pass
`format = "bibtex"`:

```{r bibtex}
bib <- project_citations(con, "GSE151530", format = "bibtex")
cat(bib)
```

This produces entries like:

```
@article{Lichun2019,
  author  = {Lichun Ma, Maria O Hernandez, ...},
  title   = {Tumor Cell Biodiversity Drives Microenvironmental ...},
  journal = {Cancer cell},
  year    = {2019},
  doi     = {10.1016/j.ccell.2019.08.007},
  pmid    = {31588021}
}
```

To save to a `.bib` file:

```{r save-bibtex}
writeLines(bib, "GSE151530.bib")
```


## Growth of scRNA-seq and snRNA-seq datasets

`assay_projects()` retrieves one row per project for each assay type,
with the publication month. We can then aggregate and plot yearly trends
with `dplyr` and `ggplot2`:

```{r sc-growth, fig.width=7, fig.height=5}
library(dplyr)
library(ggplot2)

sc_projects <- assay_projects(con,
  assay     = c("scRNA-seq", "snRNA-seq"),
  year_from = 2015,
  year_to   = 2025
)

sc_growth <- sc_projects |>
  mutate(year = as.integer(format(published, "%Y"))) |>
  count(assay_type, year, name = "n_projects")

ggplot(sc_growth, aes(x = year, y = n_projects, fill = assay_type)) +
  geom_col(position = "dodge") +
  scale_fill_manual(
    values = c("scRNA-seq" = "#3C5488", "snRNA-seq" = "#E64B35")
  ) +
  labs(
    x = "Year",
    y = "Number of GEO projects",
    fill = NULL,
    title = "Growth of scRNA-seq and snRNA-seq datasets on GEO"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```


## Finding PBMC scRNA-seq datasets with BAM files

`geo_search()` searches GEO series free-text (title, summary, overall
design) and returns the linked SRA accession. `check_sra_bam()` then
checks whether the original submission to SRA/NCBI included aligned BAM
files — useful when you want to skip alignment and work directly from
pre-aligned reads.

```{r pbmc-search}
library(dplyr)

pbmc <- geo_search(con, "PBMC scRNA-seq", limit = 10000)
```

Export the full catalogue to CSV:

```{r pbmc-csv}
write.csv(pbmc, "pbmc_scrna_geo.csv", row.names = FALSE)
```

Now check which of those studies have BAM files deposited in SRA:

```{r pbmc-bam}
pbmc_with_sra <- pbmc |> filter(!is.na(sra_accession))

bam_status <- check_sra_bam(con, pbmc_with_sra$sra_accession)

pbmc_bam <- pbmc_with_sra |>
  left_join(bam_status, by = "sra_accession") |>
  filter(has_bam_sra) |>
  arrange(desc(published))

cat(nrow(pbmc_bam), "PBMC scRNA-seq studies with BAM in SRA\n")
```

Fetch publication DOIs and render an interactive table:

```{r pbmc-bam-table}
library(DT)

pubs <- publications(con, pbmc_bam$accession)

pbmc_bam |>
  left_join(pubs |> select(accession, doi) |> distinct(),
    by = "accession"
  ) |>
  mutate(
    GEO = sprintf('<a href="https://seqout.org/p/%s">%s</a>', accession, accession),
    SRA = sprintf('<a href="https://seqout.org/p/%s">%s</a>', sra_accession, sra_accession),
    Title = ifelse(
      !is.na(doi),
      sprintf('<a href="https://doi.org/%s">%s</a>', doi, title),
      title
    )
  ) |>
  select(GEO, SRA, Title, published) |>
  datatable(
    escape = FALSE, rownames = FALSE,
    colnames = c("GEO", "SRA", "Title", "Published"),
    options = list(pageLength = 25, dom = "tip")
  )
```


## Cumulative project deposits across databases

For a broader view, `growth_stats()` returns monthly project counts
per database. We can compute cumulative totals:

```{r cumulative-growth, fig.width=7, fig.height=5}
growth <- growth_stats(con, mode = "projects") |>
  filter(db %in% c("geo", "sra")) |>
  arrange(db, month) |>
  group_by(db) |>
  mutate(cumulative = cumsum(count)) |>
  ungroup()

ggplot(growth, aes(x = month, y = cumulative, colour = db)) +
  geom_line(linewidth = 0.6) +
  scale_y_continuous(labels = scales::comma) +
  scale_colour_manual(
    values = c(geo = "#E64B35", sra = "#4DBBD5"),
    labels = c(geo = "GEO", sra = "SRA")
  ) +
  labs(
    x = NULL,
    y = "Cumulative projects",
    colour = "Database",
    title = "Cumulative number of projects deposited on GEO and SRA"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```


## Using SQL and dplyr directly

All the above functions run on DuckDB under the hood. You can also query
the tables directly with SQL:

```{r sql}
query(con, "
  SELECT dominant_scientific_name AS organism, count(*) AS n_projects
  FROM unified_metadata
  WHERE dominant_scientific_name IS NOT NULL
  GROUP BY organism
  ORDER BY n_projects DESC
  LIMIT 10
")
```

Or use `dplyr`/`dbplyr` for lazy evaluation:

```{r dplyr}
library(dbplyr)

tbl(con$db, "unified_metadata") |>
  filter(
    dominant_scientific_name == "Homo sapiens",
    country == "Japan"
  ) |>
  count(dominant_assay_l2, sort = TRUE) |>
  head(10) |>
  collect()
```


## Cache tables locally

For repeated analysis, we can materialise a remote view into local DuckDB storage
to avoid repeated HTTP requests:

```{r cache}
cache_table(con, "unified_metadata")

# Subsequent queries hit local storage
query(con, "SELECT count(*) AS n FROM unified_metadata_local")

# Clean up
clear_cache(con)
```


## Close connection

```{r close}
seqout_close(con)
```
