---
title: "Searching"
description: >
  Search seven repositories with one function: the filters, the sort order
  and the cost of a large result set.
output:
  rmarkdown::html_vignette:
    self_contained: true
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Searching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
api_reachable <- tryCatch(
  curl::curl_fetch_memory(
    "https://seqout.org/api/project/GSE151530",
    curl::new_handle(timeout = 30)
  )$status_code == 200L,
  error = function(e) FALSE
)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = api_reachable
)
```

`seqout_search()` searches seven public repositories at the same time: GEO,
SRA, ArrayExpress, ENA, GSA, DRA and GEA. It is the only search function.

Give it a query, filters, or both.

```{r first-search}
library(seqout)

seqout_search("liver cancer scRNA", limit = 5)
```

A query alone searches the title, the summary and the design of every study.
The server ranks the results.

A query is not necessary if you give at least one filter:

```{r filters-only}
seqout_search(organism = "Mus musculus", assay_l1 = "Transcriptomic", limit = 5)
```

## Filters

Give filters by name. The example below finds human RNA-seq studies about the
liver.

```{r filters}
seqout_search(
  "liver",
  organism         = "Homo sapiens",
  library_strategy = "RNA-Seq",
  limit            = 5
)
```

These are the filter names:

| Filter | What it selects |
|---|---|
| `organism` | The scientific name, for example `"Homo sapiens"`. |
| `library_strategy` | The library strategy, for example `"RNA-Seq"`. |
| `library_source` | The library source. |
| `platform` | The sequencing platform. |
| `instrument_model` | The instrument model. |
| `db`, `source` | One archive. Both names do the same thing. |
| `country` | The country name. |
| `center` | The sequencing center. |
| `journal` | The journal name. |
| `year_from`, `year_to` | The year bounds. |
| `date_from`, `date_to` | The date bounds, as `yyyy-mm-dd`. |
| `multi_platform` | Studies that use more than one platform. |
| `assay_l1`, `assay_l2` | The assay class, broad or specific. |
| `geo_country_code_iso2` | The country, as a two-letter code. |
| `geo_lat`, `geo_lng`, `geo_radius_km` | A circle on the map. |

The package refuses a name that is not in this table. The error message shows
you the names that are correct.

```{r bad-filter, error = TRUE}
seqout_search("liver", min_samples = 10)
```

### The archives

`db` and `source` accept `"geo"`, `"sra"`, `"arrayexpress"`, `"ena"`,
`"gsa"`, `"dra"` and `"gea"`.

```{r one-archive}
seqout_search("hepatocellular carcinoma", db = "geo", limit = 5)
```

### The assay class

Each study has an assay class with two levels. Level 1 is broad. Level 2 is
specific.

| `assay_l1` | Example of `assay_l2` |
|---|---|
| Transcriptomic | RNA-seq |
| Genomic Sequencing | WGS |
| Chromatin Binding | ChIP-seq |
| Microarray | Expression Microarray |
| DNA Methylation | Bisulfite-seq |

The match is exact. Use the capital letters that the table shows.

```{r assay}
seqout_search("liver", assay_l2 = "scRNA-seq", limit = 5)
```

### The dates

`year_from` and `year_to` take whole years.

```{r years}
seqout_search("liver cancer", organism = "Homo sapiens", year_from = 2023, limit = 5)
```

`date_from` and `date_to` take a day. Write the day as `yyyy-mm-dd`. These two
filters bound `updated_at`. `updated_at` is the day when seqout last saw a
change to the record.

```{r dates}
seqout_search("liver cancer",
  date_from = "2024-01-01", date_to = "2024-01-31", limit = 5
)
```

## The order of the results

The default order is relevance. Change it with `sortby`.

```{r sortby}
seqout_search("hepatocellular carcinoma", sortby = "citations", limit = 5)
```

`sortby` accepts `"citations"`, `"journal"` and `"year"`. `order` accepts
`"desc"`, which is the default, and `"asc"`.

## How many rows you get

A search returns every match. The server sends 200 rows in one page, so a large
result set costs more than one request.

```{r all-rows}
result <- seqout_search("hepatocellular carcinoma scRNA")
nrow(result)
```

Set `limit` when a sample is sufficient. A limit of 50 costs one request.

```{r limit}
nrow(seqout_search("hepatocellular carcinoma scRNA", limit = 50))
```

## Two more functions

`search_correction()` gives you the correction that the server suggests for a
query with a spelling mistake. It costs one request.

```{r correction}
found <- search_correction("hepatocelular carcinima")
found$correction$corrected_query
```

`bulk_search()` runs one search for each query in a vector. It gives you a
named list of tibbles.

```{r bulk}
many <- bulk_search(
  c("hepatocellular carcinoma", "pancreatic ductal adenocarcinoma"),
  db = "geo", limit = 10
)
vapply(many, nrow, integer(1))
```

`bulk_search()` runs the searches one after the other. It also replaces a
search that fails with an empty tibble. Give `limit` to keep the cost low.

## From a result to a study

A search gives you accessions. Use `project()` to read the full record of one
of them.

```{r project}
top <- seqout_search("hepatocellular carcinoma", sortby = "citations", limit = 1)
project(top$accession[1], transpose = TRUE)
```

`dataset()` does more. It starts from one accession and finds the samples, the
experiments, the runs and the publications.

```{r dataset}
d <- dataset(top$accession[1])
c(samples = nrow(d$samples), runs = nrow(d$runs))
```

## The Parquet backend

`seqout_search()` reads the REST API. It does not read the Parquet dump.

The Parquet dump answers SQL. Use SQL for a filter or a count over the full
index: it does more than a search argument, and it is faster for a large scan.
Open a Parquet connection, then write the query.

```{r parquet, eval = FALSE}
con <- seqout_connect("parquet")

query("
  SELECT dominant_scientific_name AS organism, count(*) AS n
  FROM unified_metadata
  WHERE n_samples >= 10
  GROUP BY organism
  ORDER BY n DESC
  LIMIT 10
", con = con)
```

The package tells you when you give a Parquet connection to `seqout_search()`.

```{r parquet-error, eval = FALSE}
seqout_search("liver", con = con)
#> Error: `seqout_search()` reads the REST API; this connection is Parquet.
#> ℹ Drop `con` to search, or use `query()` to write SQL over the dump.
```
