---
title: "Searching"
description: >
  Search seven repositories with one function: filters, sort order and the
  cost of a large result set.
output:
  rmarkdown::html_vignette:
    self_contained: true
    mathjax: null
vignette: >
  %\VignetteIndexEntry{Searching}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
seqout::seqout_default(NULL)
```

`seqout_search()` searches GEO, SRA, ArrayExpress, ENA, GSA, DRA and GEA at
once:

```{r first-search}
library(seqout)

seqout_search("liver cancer scRNA", limit = 5)
```

A query matches the title, summary and design of every study. We can optionally apply filters to the query:

```{r filters-only}
seqout_search(organism = "Mus musculus", assay_l1 = "Transcriptomic", limit = 5)
```

## Filters

```{r filters}
seqout_search(
  "liver",
  organism         = "Homo sapiens",
  library_strategy = "RNA-Seq",
  limit            = 5
)
```



Limiting search to a database:

```{r one-archive}
seqout_search("hepatocellular carcinoma", db = "geo", limit = 5)
```

Limiting search to an assay:

```{r assay}
seqout_search("liver", assay_l2 = "scRNA-seq", limit = 5)
```

Limiting search by year:

```{r years}
seqout_search("liver cancer", organism = "Homo sapiens", year_from = 2023, limit = 5)
```


```{r dates}
seqout_search("liver cancer",
  date_from = "2024-01-01", date_to = "2024-01-31", limit = 5
)
```

## Sort order

Results come back by relevance unless `sortby` is set. It takes
`"citations"`, `"journal"` and `"year"`; `order` takes `"desc"` (the default)
and `"asc"`.

```{r sortby}
seqout_search("hepatocellular carcinoma", sortby = "citations", limit = 5)
```


## How many rows you get

A search returns every match. The server sends 200 rows in one page, so a large
result set costs more than one request.

```{r all-rows}
result <- seqout_search("hepatocellular carcinoma scRNA")
nrow(result)
```

Set `limit` when a sample is sufficient. A limit of 50 costs one request.

```{r limit}
nrow(seqout_search("hepatocellular carcinoma scRNA", limit = 50))
```

## From a result to a study

A search gives you accessions. Use `project()` to read the full record of one
of them.

```{r project}
top <- seqout_search("hepatocellular carcinoma", sortby = "citations", limit = 1)
project(top$accession[1], transpose = TRUE)
```

`seqout_get()` does more. It starts from one accession and finds the samples, the
experiments, the runs and the publications.

```{r dataset}
d <- seqout_get(top$accession[1])
c(samples = nrow(d$samples), runs = nrow(d$runs))
```

## Parquet backend

`seqout_search()` is REST-only. For a filter or count over the whole index, you can write 
SQL against the Parquet dump:

```{r parquet, eval = FALSE}
con <- seqout_connect("parquet")

query("
  SELECT dominant_scientific_name AS organism, count(*) AS n
  FROM unified_metadata
  WHERE n_samples >= 10
  GROUP BY organism
  ORDER BY n DESC
  LIMIT 10
", con = con)
```


