---
title: "Citations"
description: >
  Get BibTeX for the papers behind a dataset, for one accession or for a
  bibliography.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A dataset is evidence, and evidence needs a citation. The archives already
record which paper each dataset belongs to, so `Citations()` assembles the
BibTeX entry from that record rather than making you find the paper by hand.

```{r setup}
library(seqout)
```

## One dataset

`Citations()` takes a project accession and returns text.

```{r one}
cat(Citations("GSE151530"))
```

The key is the first author's family name and the year. Write it to a file the
same way you write any text.

```{r write, eval = FALSE}
writeLines(Citations("GSE151530"), "GSE151530.bib")
```

## The paper, or every paper

`type = "original"`, the default, gives the paper that the submitters wrote.
`type = "all"` adds the papers that reanalysed the data afterwards. The second
list is the one to read when you want to know what a dataset has been used for.

```{r all}
cat(Citations("GSE168652", type = "all"))
```

That study has one original paper and one later reanalysis.

## A dataset with no paper

Many datasets are deposited without a publication, and a great many are
published later. `Citations()` returns `character(0)` for those, rather than an
error, so a loop does not stop at the first gap.

```{r none}
length(Citations("CRA027437"))
```

## A bibliography

Because the result is plain text, several accessions concatenate.

```{r many}
accessions <- c("GSE151530", "GSE168652", "SRP267291", "CRA027437")
bib <- unlist(lapply(accessions, Citations))

length(bib)
```

Four accessions, three entries: the one without a paper contributed nothing,
and `unlist()` dropped it.

```{r many-write, eval = FALSE}
writeLines(bib, "study-bibliography.bib")
```

Test first when you need to know which accession gave which entry.

```{r many-named}
vapply(accessions, function(a) length(Citations(a)) > 0, logical(1))
```

## The same papers as data

BibTeX is for a manuscript. When you want to sort, count or filter, read the
publication records instead: `$pubs` carries the same papers as a tibble, with
the journal and the citation count.

```{r pubs}
d <- SeqoutGet("GSE168652")
d$pubs[, c("pmid", "title", "journal", "citation_count")]
```

[Metadata](metadata.html#publications) describes that field. To go the other
way, from a paper or an author to the datasets, [Search](search.html) covers
`Paper()` and `Author()`.

## The REST API only

`Citations()` reads the REST API, and says so on a Parquet connection.

```{r parquet, error = TRUE}
con <- SeqoutConnect("parquet", quiet = TRUE)
Citations("GSE151530", con = con)
```

```{r close, include = FALSE}
SeqoutClose(con)
```

The dump could answer a version of this, and deliberately does not. It has no
record of the papers that reanalysed a dataset, and only about a third of its
publication rows carry a date, so an entry built from it would be missing its
year without saying so. A citation is not worth getting nearly right.
