---
title: "Retrieve Dataset Citations"
description: >
  Learn how to retrieve BibTeX citations for publication-linked datasets, compile bibliographies, and access publication metadata.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`Citations()` returns BibTeX entries for the papers linked to a dataset.

```{r setup}
library(seqout)
```

## Retrieve a single citation

Pass a project accession to `Citations()`:

```{r one}
cat(Citations("GSE151530"))
```

The citation key is `[FirstAuthorFamilyName][Year]`.

Write it to a file with `writeLines()`:

```{r write, eval = FALSE}
writeLines(Citations("GSE151530"), "GSE151530.bib")
```

## Distinguish original papers from reanalyses

By default `Citations()` returns the original dataset paper (`type = "original"`).
`type = "all"` adds reanalyses:

```{r all}
cat(Citations("GSE168652", type = "all"))
```


## Handle datasets without publications

Many datasets are deposited before publication, and some have no linked paper.
`Citations()` then returns `character(0)`:

```{r none}
length(Citations("CRA027437"))
```

## Compile a bibliography

Given several accessions, `Citations()` returns one character vector, each entry
named for its accession:

```{r many}
accessions <- c("GSE151530", "GSE168652", "SRP267291", "CRA027437")
bib <- Citations(accessions)

length(bib)
```

An accession without a paper adds no entry. Write the bibliography with
`writeLines()`:

```{r many-write, eval = FALSE}
writeLines(bib, "study-bibliography.bib")
```

The names show which accessions produced a citation:

```{r many-named}
table(names(bib))
```

## Access publication metadata as data frames

`SeqoutGet()` returns publication records as a data frame in `pubs`:

```{r pubs}
d <- SeqoutGet("GSE168652")
d$pubs[, c("pmid", "title", "journal", "citation_count")]
```

See [Metadata](metadata.html#mapped-publications) for more, and [Search](search.html) to find datasets by author or paper.

## Backend support

`Citations()` queries the REST API. A Parquet connection raises an error:

```{r parquet, error = TRUE}
con <- SeqoutConnect("parquet", quiet = TRUE)
Citations("GSE151530", con = con)
```

```{r close, include = FALSE}
SeqoutClose(con)
```

> [!NOTE]
> The dump lacks reanalysis publications and many release years, so `Citations()` needs the REST API.
