---
title: "Retrieve Dataset Citations"
description: >
  Learn how to retrieve BibTeX citations for publication-linked datasets, compile bibliographies, and access publication metadata.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

To ensure scientific reproducibility, you should cite the original publications associated with any datasets used in your research. Public repositories record the publications linked to each dataset. 

The `seqout` package provides the `Citations()` function to automatically retrieve and format these publication records as BibTeX entries, saving you from searching for literature details manually.

```{r setup}
library(seqout)
```

## Retrieve a single citation

To fetch the BibTeX citation for a specific project accession, call `Citations()` and pass the accession ID:

```{r one}
cat(Citations("GSE151530"))
```

The generated citation key uses the format `[FirstAuthorFamilyName][Year]`. 

To save the BibTeX citation directly to a file, use the `writeLines()` function:

```{r write, eval = FALSE}
writeLines(Citations("GSE151530"), "GSE151530.bib")
```

## Distinguish original papers from reanalyses

By default, `Citations()` retrieves the paper published by the dataset's original authors (`type = "original"`). 

To include both the original publication and subsequent publications that reanalyzed the dataset, set `type = "all"`. This allows you to explore how a dataset has been used across different studies:

```{r all}
cat(Citations("GSE168652", type = "all"))
```

In this example, the study is linked to one original paper and one reanalysis paper.

## Handle datasets without publications

Many datasets are deposited before publication, and some may not have an associated paper. 

For datasets without a recorded publication, `Citations()` returns an empty character vector (`character(0)`) rather than raising an error. This behavior ensures that batch queries and loops do not terminate when encountering a dataset without a citation:

```{r none}
length(Citations("CRA027437"))
```

## Compile a bibliography

Because `Citations()` returns plain text, you can concatenate citations from multiple accessions into a single bibliography file:

```{r many}
accessions <- c("GSE151530", "GSE168652", "SRP267291", "CRA027437")
bib <- unlist(lapply(accessions, Citations))

length(bib)
```

In this example, querying four accessions returns three bibliography entries. The accession without a recorded publication returned `character(0)`, which `unlist()` automatically dropped.

To write the compiled bibliography to disk, call `writeLines()`:

```{r many-write, eval = FALSE}
writeLines(bib, "study-bibliography.bib")
```

If you need to verify which accessions successfully returned a citation, check their return lengths first:

```{r many-named}
vapply(accessions, function(a) length(Citations(a)) > 0, logical(1))
```

## Access publication metadata as data frames

While BibTeX is suitable for rendering bibliographies, you may want to filter or count publications programmatically. 

To retrieve publication records as a structured data frame, access the `pubs` field of the dataset object returned by `SeqoutGet()`. The returned table includes publication details like the journal name and citation count:

```{r pubs}
d <- SeqoutGet("GSE168652")
d$pubs[, c("pmid", "title", "journal", "citation_count")]
```

For more details on dataset metadata, see [Metadata](metadata.html#publications). To search for datasets using a specific author name or publication identifier, see [Search](search.html).

## Backend support

The `Citations()` function queries the REST API and is not supported by the Parquet backend. Attempting to run `Citations()` on a Parquet connection raises an error:

```{r parquet, error = TRUE}
con <- SeqoutConnect("parquet", quiet = TRUE)
Citations("GSE151530", con = con)
```

```{r close, include = FALSE}
SeqoutClose(con)
```

> [!NOTE]
> The database dump does not store citation data because it lacks records for reanalysis publications, and many publication records do not contain release years. Restricting `Citations()` to the REST API ensures that you always receive complete and accurate citation metadata.
