---
title: "Counts"
description: >
  Find the counts matrices of a GEO study, read one, add the sample metadata,
  and make a Seurat object.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

GEO keeps processed data in supplementary files. GEO gives these files no
schema. Each submitter selects their own file names, formats and layout.

Seqout reads these deposits retroactively. It looks at the files of a study
after the submitter sent them, puts related files into groups, and reads each
group as a matrix. Thus a study from 2013 and a study from this year go through
the same three steps:

1. `SeqoutCounts()` tells you which matrices the study has.
2. `SeqoutMatrix()` reads one of them.
3. `Seqout2Seurat()` makes a Seurat object from it.

Because the grouping is retroactive, it is an interpretation of the files and
not a fact from GEO. Look at the table from step 1 before you trust step 2.

## Before you start

The readers need `Matrix` for the sparse formats. They need `hdf5r` for `.h5`
and `.h5ad`. Seqout suggests both packages and does not require them, thus a
plain installation stays small.

```{r install, eval = FALSE}
install.packages(c("Matrix", "hdf5r"))
```

```{r setup}
library(seqout)

ACCESSION <- "GSE297547" # PBMCs from myasthenia gravis patients and controls
```

## Step 1: see what the study has

`SeqoutCounts()` asks the API for the file list of a GEO accession. It then
returns one row for each matrix that it can read. It downloads no data, because
the files of one series can be tens of gigabytes.

```{r counts-table}
counts <- SeqoutCounts(ACCESSION)

nrow(counts)
head(counts[, c("unit", "sample", "format", "preferred", "has_metadata")], 8)
```

Each row is a **unit**: the smallest group of files that Seqout can read as one
matrix. A unit is a 10x triplet (`matrix.mtx` with `barcodes.tsv` and
`features.tsv`), a CellRanger `.h5`, an `.h5ad`, an `.rds`, or a delimited
table.

The columns tell you this:

| Column | What it tells you |
| --- | --- |
| `unit` | The label that selects this exact matrix |
| `sample` | The GSM, when Seqout can assign one |
| `format` | The reader that Seqout uses |
| `preferred` | Seqout takes this unit when you ask for the sample |
| `has_metadata` | Cell annotation is probable |
| `n_files`, `files` | How many files the unit has, and their names |

One sample frequently has more than one unit. The label of each unit then gets
a suffix, for example `GSM8994520:h5ad`.

```{r formats}
table(counts$format)
```

### How Seqout selects the preferred unit

Submitters frequently deposit the same counts two times. A 10x `.h5` beside the
mtx triplet that made it is a common example. The table shows both units.

`preferred` marks the unit that `SeqoutMatrix()` takes when you name a GSM and
not a unit label. Seqout ranks the units of a sample in this sequence:

1. The assay that you asked for.
2. Filtered output before raw output.
3. Format: 10x MatrixMarket, then 10x HDF5, then `.h5ad`, then `.rds`, then
   tar, then a delimited table.

A unit that is not preferred stays available. Give its `unit` label to
`SeqoutMatrix()`.

### What `has_metadata` tells you

`has_metadata` is a signal from discovery and not a check of content. It is
`TRUE` in two conditions:

* The unit is an `.h5ad` or an `.rds`. These formats can hold cell annotation
  in the container.
* Seqout found a related file whose name shows cell annotation, for example
  `metadata`, `annotation`, `celltype` or `obs.csv`.

Seqout attaches a file that has a sample name to the units of that sample. It
attaches a file that has no sample name to all the units of the series. Seqout
reads the annotation only when `SeqoutMatrix()` reads the unit.

## Step 2: read one matrix

`SeqoutMatrix()` downloads the files of one unit and reads them. It keeps the
files in a cache directory below `tools::R_user_dir("seqout", "cache")`. The
cache stays after the session ends, thus a second read of the same unit
downloads nothing.

```{r read-rna}
rna <- SeqoutMatrix(counts, sample = "GSM8994520")
rna
```

A `seqout_matrix` is a list. `X` holds the counts, `obs` holds the observation
annotation, and `var` holds the feature annotation.

`X` is features by observations: genes by cells for single-cell data, and genes
by biological samples for bulk data. This is the orientation that Seurat and
most Bioconductor classes expect, thus you do not transpose it.

```{r shape}
dim(rna$X)
class(rna$X)
```

### Single cell or bulk

Seqout decides `kind` from evidence and tells you the evidence. Row labels that
look like barcodes mean single cell. Row labels that are GSM accessions mean
bulk. If the labels give no answer, Seqout counts the observations: 32 or fewer
is bulk, 400 or more is single cell.

Between those two numbers the data can be a Smart-seq plate or a small bulk
study. Seqout reports `unknown` and does not guess. Read `evidence` when the
answer is important to you.

```{r evidence}
rna$kind
rna$evidence
```

### The cell annotation

`obs` holds the annotation that the submitter deposited for each cell. For an
`.h5ad` or an `.rds` it comes from the container. For a 10x unit it comes from
a related file, which Seqout joins on the cell label.

```{r obs-rna}
lapply(rna$obs, function(x) head(unique(as.character(x)), 3))
```

Column names change from submitter to submitter. This submitter named the
columns `type1`, `type2` and `type3`. These names contain no vocabulary that a
program can match, thus Seqout does not tell you which column holds the cell
type. The columns stay available to you: `type3` is the donor code and `type1`
is the donor group.

Examine the gene names before you plan an analysis. This submitter made the
gene names anonymous. Marker-based cell typing is therefore not possible for
this study.

```{r gene-names}
head(rownames(rna$var), 5)
```

## Read a different assay

This is a CITE-seq study. Each sample has two matrices: gene expression and an
antibody panel. Both are `.h5ad` files, thus the format separates neither one,
and the antibody file comes first in the file list.

For this reason `SeqoutCounts()` uses `assay = "rna"` as the default. The other
values are `"adt"` (antibody-derived tags), `"hto"` (hashtag oligos), `"atac"`,
and `NULL` for no preference.

Seqout identifies the assay from parts of the file name, for example `gex`,
`antibody`, `hashtag` and `peak`. It then uses your assay as the first rule
when it ranks the units. For 10x files it also keeps the related feature class:
Gene Expression for RNA, Antibody Capture for ADT and HTO, and Peaks for ATAC.
To override that filter, set `feature_type`.

```{r read-adt}
adt <- SeqoutCounts(ACCESSION, assay = "adt")
adt_m <- SeqoutMatrix(adt, sample = "GSM8994520")

dim(adt_m$X)
rownames(adt_m$var)
```

The antibody unit of this study has the cell types in `obs`.

```{r celltype}
head(sort(table(adt_m$obs$Cell_Type_Experimental), decreasing = TRUE))
```

## Add the study metadata

The counts tell you what the experiment measured. They tell you nothing about
the donors, the condition, or the paper. That data is in the API.

```{r project}
d <- SeqoutGet(ACCESSION)

d$meta$title
d$meta$overall_design
```

`d$samples` is the design of the experiment. It has one row for each GSM, and
the keys of the submitter as columns.

```{r design}
design <- d$samples[, c("accession", "title", "age", "Sex", "tissue")]
head(design)
```

Each cell of a unit belongs to one GSM, which has one row in that table.

```{r join}
design[design$accession == rna$accession, ]
```

Compare the two sources before you trust either one. `type1` and `type3` are
the per-cell labels of the submitter. `design` is what GEO holds.

```{r cross-check}
picked <- c("GSM8994520", "GSM8994523", "GSM8994533")

cells <- do.call(rbind, lapply(picked, function(a) {
  o <- SeqoutMatrix(counts, sample = a)$obs
  data.frame(accession = a, per_cell = unique(o$type1), donor = unique(o$type3))
}))
merge(cells, design[, c("accession", "title")], by = "accession")
```

The per-cell labels and the donor codes agree.

## Step 3: make a Seurat object

`Seqout2Seurat()` builds the object. Give it a `seqout_matrix`, or give it one
GSM accession when you want one sample and nothing more. The arguments of
`CreateSeuratObject()`, for example `project`, `min.cells` and `min.features`,
go through to that function.

```{r seurat-one, message = FALSE, warning = FALSE}
library(Seurat)

Seqout2Seurat(SeqoutMatrix(counts, sample = picked[1]), project = picked[1])
```

Seqout does not transpose the matrix, because `X` already has the orientation
that Seurat expects. The cell annotation from `obs` becomes `meta.data`.

For more than one sample, add the covariates of each sample after you make its
object.

```{r seurat-build, message = FALSE, warning = FALSE}
library(Matrix)
set.seed(1)

n_cells <- 1500

objs <- lapply(picked, function(accession) {
  m <- SeqoutMatrix(counts, sample = accession)
  keep <- sample(ncol(m$X), min(n_cells, ncol(m$X)))
  m$X <- m$X[, keep, drop = FALSE]
  m$obs <- m$obs[keep, , drop = FALSE]
  covariates <- design[design$accession == accession, ]

  obj <- Seqout2Seurat(m, project = accession, min.cells = 3, min.features = 200)
  obj$sample <- accession
  obj$donor <- sub(".*from ([A-Z]+[0-9]+).*", "\\1", covariates$title)
  obj$group <- ifelse(grepl("^MG", obj$donor), "MG", "control")
  obj$Sex <- covariates$Sex
  obj$age <- as.integer(covariates$age)
  obj
})
```

The covariates join the cell annotation in `meta.data`.

```{r seurat-merge, message = FALSE, warning = FALSE}
pbmc <- merge(objs[[1]], objs[-1],
  add.cell.ids = vapply(objs, function(o) unique(o$donor), character(1))
)
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

table(pbmc$donor, pbmc$group)
```

## An example analysis

From this point the work is Seurat and not Seqout.

```{r sctransform, message = FALSE, warning = FALSE}
pbmc <- SCTransform(pbmc, vars.to.regress = "percent.mt", verbose = FALSE)
pbmc <- RunPCA(pbmc, npcs = 30, verbose = FALSE)
pbmc <- RunUMAP(pbmc, dims = 1:30, verbose = FALSE)
pbmc <- FindNeighbors(pbmc, dims = 1:30, verbose = FALSE)
pbmc <- FindClusters(pbmc, resolution = 0.4, verbose = FALSE)

pbmc
```

```{r umap-clusters, fig.width = 6, fig.height = 5, message = FALSE, warning = FALSE}
DimPlot(pbmc, label = TRUE) + ggplot2::ggtitle("Clusters")
```

```{r umap-metadata, fig.width = 10, fig.height = 4.5, message = FALSE, warning = FALSE}
DimPlot(pbmc, group.by = c("donor", "group", "Sex"), ncol = 3)
```

```{r composition}
round(prop.table(table(pbmc$seurat_clusters, pbmc$group), margin = 2), 3)
```

`QuickAnnotation()` puts a label on each cluster from marker sets that you
supply. It scores each cell as the mean expression of the genes of a set,
calculates the mean of those scores in each cluster, and takes the highest
score. The scores of different sets have different scales. Read
`attr(labels, "scores")` before you trust a label.

The gene names of this study are anonymous, thus the code below is an example
and not a result.

```{r annotation, eval = FALSE}
markers <- list(
  t_cell = c("CD3D", "CD3E", "IL7R"),
  b_cell = c("MS4A1", "CD79A"),
  myeloid = c("LYZ", "CD14", "FCGR3A")
)

expr <- SeuratObject::LayerData(JoinLayers(pbmc), layer = "data")
labels <- QuickAnnotation(expr, pbmc$seurat_clusters, markers)
pbmc$celltype <- labels[as.character(pbmc$seurat_clusters)]
```

## More than one matrix

`Matrices()` reads each preferred unit and returns a list, with the unit labels
as names. It gets all the files in one call, because one file at a time is the
slow way to download a series of 19 samples.

```{r whole-study, eval = FALSE}
mats <- Matrices(counts)
```

Seqout skips a unit that it cannot read and gives a warning. One bad file thus
does not stop the full call. Compare the length of the result with the number
of preferred rows in the counts table.

`BindCounts()` joins more than one matrix into one matrix, on the features that
they have in common. It puts a label before each cell name, and it can set a
maximum number of cells for each matrix.

```{r bind, eval = FALSE}
merged <- BindCounts(Matrices(counts, sample = picked),
  labels = picked, max_cells = 1200
)
dim(merged)
```

`SeqoutCountsFiles()` downloads the files of a unit and reads none of them. Use
it when you want to read the files with your own tools.

```{r files, eval = FALSE}
paths <- SeqoutCountsFiles(counts, sample = "GSM8994520")
```

## What it costs

The download is the expensive step. One series can be tens of gigabytes. The
counts table costs one API call, thus you can look at a study before you commit
to it. Read the table first, and take only the units that you need.
