---
title: "Counts"
description: >
  Turn a GEO accession into a counts matrix, join the sample metadata, and
  carry both into Seurat.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The readers need `Matrix` for the sparse formats and `hdf5r` for `.h5` and
`.h5ad`. Both are suggested and not required, so a plain install stays light.

```{r install, eval = FALSE}
install.packages(c("Matrix", "hdf5r"))
```

```{r setup}
library(seqout)

ACCESSION <- "GSE297547" # PBMCs from myasthenia gravis patients and controls
```

## The manifest comes first

`SeqoutCounts()` resolves the supplementary files of a GEO accession and
groups them into units. It downloads nothing, because a GEO payload can be tens
of gigabytes.

A unit is a group of files that read as one matrix: a 10x triplet (`matrix.mtx`
with `barcodes.tsv` and `features.tsv`), a CellRanger `.h5`, an `.h5ad`, an
`.rds`, or a delimited table.

```{r manifest}
counts <- SeqoutCounts(ACCESSION)
m <- Manifest(counts)

nrow(m)
head(m[, c("unit", "sample", "format", "preferred", "has_metadata")], 8)
```

`preferred` marks the unit that `SeqoutMatrix()` takes for each sample. A
submitter often deposits the same counts twice, for example a 10x `.h5` beside
the mtx triplet that it was built from. The manifest lists both. The unit that
loses stays reachable by its label, so nothing is hidden from you.

`has_metadata` says whether per-cell annotation is available at all, either as
a sidecar file or inside the container. An `.h5ad` and an `.rds` always carry
their own.

```{r formats}
table(m$format)
```

## Read one unit

`SeqoutMatrix()` downloads the files of one unit, caches them under
`tools::R_user_dir("seqout", "cache")`, and reads them. The cache lasts across
sessions, so a second read costs nothing.

```{r read-rna}
rna <- SeqoutMatrix(counts, sample = "GSM8994520")
rna
```

A `seqout_matrix` holds the counts in `X`, the observation annotation in `obs`
and the feature annotation in `var`. `X` is in AnnData orientation: the rows
are observations and the columns are genes. For single-cell data a row is a
cell. For bulk data it is a biological sample. One type therefore covers both,
and the caller does not check which they have.

```{r shape}
dim(rna$X)
class(rna$X)
```

`CountsMatrix()` gives the other orientation, features by observations, which
is what Seurat and most Bioconductor classes expect. It is also exported as
`genexcell_counts()`. `cellxgene_counts()` returns `X` unchanged.

```{r counts-matrix}
dim(CountsMatrix(rna))
```

### Single cell or bulk

`kind` is decided on evidence. Row labels that look like barcodes mean single
cell. Row labels that are GSM accessions, or an observation count no larger
than the number of samples in the series, mean bulk. A dense table with a few
hundred columns could be a Smart-seq plate or a large bulk study, and in that
band `kind` reports `unknown`. Read `evidence` when the answer matters.

```{r evidence}
rna$kind
rna$evidence
```

## Cell-level annotation

`obs` carries whatever per-cell annotation the submitter deposited. For an
`.h5ad` or an `.rds` it comes from inside the container. For a 10x unit it
comes from a sidecar file, joined on the cell label.

```{r obs-rna}
lapply(rna$obs, function(x) head(unique(as.character(x)), 3))
```

Column names differ widely between submitters. This one named their annotation
`type1`, `type2` and `type3`, which carries no vocabulary that a matcher could
key on. Nothing here therefore claims that a particular column is the cell
type. The columns are still there and still usable: `type3` is the donor code
and `type1` is the donor group, both recorded per cell.

Check the gene names before you plan an analysis. This submitter deposited them
anonymised, which puts marker-based cell typing out of reach for this study.

```{r gene-names}
head(rownames(rna$var), 5)
```

## Select an assay

This is a CITE-seq study, so every sample ships two matrices: gene expression
and an antibody panel. Both are `.h5ad`, so the file type separates neither,
and the antibody file happens to sort first.

`SeqoutCounts()` defaults to `assay = "rna"` for that reason. Ask for the
panel and you get it, with its own annotation. `assay` also takes `"adt"`,
`"hto"` and `"atac"`.

```{r read-adt}
adt <- SeqoutCounts(ACCESSION, assay = "adt")
adt_m <- SeqoutMatrix(adt, sample = "GSM8994520")

dim(adt_m$X)
rownames(adt_m$var)
```

```{r celltype}
head(sort(table(adt_m$obs$Cell_Type_Experimental), decreasing = TRUE))
```

## Join the study metadata

The counts say what was measured. They say nothing about who the donors were,
what the condition was, or which paper this came from. That lives in the API.

```{r project}
d <- SeqoutGet(ACCESSION)

d$meta$title
d$meta$overall_design
```

`d$samples` is the design of the experiment, one row per GSM, with the
submitter's own keys as columns.

```{r design}
design <- d$samples[, c("accession", "title", "age", "Sex", "tissue")]
head(design)
```

Every cell of a unit belongs to one GSM, which has its own row in that table.

```{r join}
design[design$accession == rna$accession, ]
```

Check the two sources against each other before you trust either one. `type1`
and `type3` are the submitter's per-cell labels. `design` is what GEO holds.

```{r cross-check}
picked <- c("GSM8994520", "GSM8994523", "GSM8994533")

cells <- do.call(rbind, lapply(picked, function(a) {
  o <- SeqoutMatrix(counts, sample = a)$obs
  data.frame(accession = a, per_cell = unique(o$type1), donor = unique(o$type3))
}))
merge(cells, design[, c("accession", "title")], by = "accession")
```

The per-cell labels and the donor codes agree.

## Into Seurat

Each sample now has both its sample-level metadata and its cell-level
annotation, so a deeper analysis can start.

```{r seurat-build, message = FALSE, warning = FALSE}
library(Seurat)
library(Matrix)
set.seed(1)

n_cells <- 1500

objs <- lapply(picked, function(accession) {
  x <- CountsMatrix(SeqoutMatrix(counts, sample = accession))
  keep <- sample(ncol(x), min(n_cells, ncol(x)))
  covariates <- design[design$accession == accession, ]

  obj <- CreateSeuratObject(
    counts = x[, keep, drop = FALSE],
    project = accession, min.cells = 3, min.features = 200
  )
  obj$sample <- accession
  obj$donor <- sub(".*from ([A-Z]+[0-9]+).*", "\\1", covariates$title)
  obj$group <- ifelse(grepl("^MG", obj$donor), "MG", "control")
  obj$Sex <- covariates$Sex
  obj$age <- as.integer(covariates$age)
  obj
})
```

The covariates become `meta.data` on the Seurat object.

```{r seurat-merge, message = FALSE, warning = FALSE}
pbmc <- merge(objs[[1]], objs[-1],
  add.cell.ids = vapply(objs, function(o) unique(o$donor), character(1))
)
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

table(pbmc$donor, pbmc$group)
```

```{r sctransform, message = FALSE, warning = FALSE}
pbmc <- SCTransform(pbmc, vars.to.regress = "percent.mt", verbose = FALSE)
pbmc <- RunPCA(pbmc, npcs = 30, verbose = FALSE)
pbmc <- RunUMAP(pbmc, dims = 1:30, verbose = FALSE)
pbmc <- FindNeighbors(pbmc, dims = 1:30, verbose = FALSE)
pbmc <- FindClusters(pbmc, resolution = 0.4, verbose = FALSE)

pbmc
```

```{r umap-clusters, fig.width = 6, fig.height = 5, message = FALSE, warning = FALSE}
DimPlot(pbmc, label = TRUE) + ggplot2::ggtitle("Clusters")
```

```{r umap-metadata, fig.width = 10, fig.height = 4.5, message = FALSE, warning = FALSE}
DimPlot(pbmc, group.by = c("donor", "group", "Sex"), ncol = 3)
```

```{r composition}
round(prop.table(table(pbmc$seurat_clusters, pbmc$group), margin = 2), 3)
```

`QuickAnnotation()` labels the clusters from marker sets of your own. It scores
each cell as the mean expression of a set's genes, averages that within each
cluster, and takes the highest score. The scores are unscaled across sets, so
read `attr(labels, "scores")` before you trust a label. The gene names of this
study are anonymised, so the call below is an example and not a result.

```{r annotation, eval = FALSE}
markers <- list(
  t_cell = c("CD3D", "CD3E", "IL7R"),
  b_cell = c("MS4A1", "CD79A"),
  myeloid = c("LYZ", "CD14", "FCGR3A")
)

expr <- SeuratObject::LayerData(JoinLayers(pbmc), layer = "data")
labels <- QuickAnnotation(expr, pbmc$seurat_clusters, markers)
pbmc$celltype <- labels[as.character(pbmc$seurat_clusters)]
```

## The whole study

`Matrices()` reads every preferred unit and returns a list keyed by unit label.
It fetches all the files in one call, because one stream at a time is the slow
way to download a 19-sample series.

```{r whole-study, eval = FALSE}
mats <- Matrices(counts)
```

A unit that cannot be read is skipped with a warning, so one broken file does
not fail the whole call. Compare the length of the result against the manifest.

`BindCounts()` joins several matrices into one, on the features they share. It
prefixes the cell names with a label, and it can cap the cells per matrix.

```{r bind, eval = FALSE}
merged <- BindCounts(Matrices(counts, sample = picked),
  labels = picked, max_cells = 1200
)
dim(merged)
```

`SeqoutCountsFiles()` downloads the files of a unit and parses nothing. Use it
when you want to read them with your own tools.

```{r files, eval = FALSE}
paths <- SeqoutCountsFiles(counts, sample = "GSM8994520")
```

The download is the expensive part, and a series can be tens of gigabytes. Read
the manifest first, and take the units you need.
