---
title: "Read Counts Matrices"
description: >
  Find, download, and read counts matrices for a GEO study, join study metadata, and convert them to Seurat objects.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Gene Expression Omnibus (GEO) stores processed single-cell and bulk genomic datasets as supplementary files. GEO has no strict schema for those files, so authors use many formats, file names, and directory layouts.

`seqout` identifies related supplementary files, groups them into readable units, and parses those units as counts matrices.

A typical R workflow is:

1. Inspect matrix units with `SeqoutListCounts()`.
2. Read one unit with `SeqoutMatrix()` or a set of units with `Seqout2Seurat()`.
3. Join study metadata from `SeqoutGet()` when the analysis needs donor or group covariates.

> [!NOTE]
> Review the manifest to select files before downloading.

## Prerequisites

Sparse matrices need `Matrix`. CellRanger `.h5` and AnnData `.h5ad` files need `hdf5r`.

```{r install, eval = FALSE}
install.packages(c("Matrix", "hdf5r"))
```

Load `seqout` and set the GEO accession used throughout the article:

```{r setup}
library(seqout)
library(Seurat)

ACCESSION <- "GSE297547" # PBMCs from patients with myasthenia gravis and healthy controls
```

## Step 1: Identify available datasets

`SeqoutListCounts()` returns the readable matrix units for a GEO study or sample. It queries file metadata and downloads no matrix files.

```{r counts-table}
counts <- SeqoutListCounts(ACCESSION)

nrow(counts)
head(counts[, c("unit", "sample", "format", "preferred", "has_metadata")], 8)
```

Each row is a **unit**, the smallest file group that `seqout` can parse as one matrix. Common units include:

* 10x Genomics triplets (`matrix.mtx`, `barcodes.tsv`, and `features.tsv`)
* CellRanger HDF5 files (`.h5`)
* AnnData objects (`.h5ad`)
* R serialized datasets (`.rds`)
* Delimited tables (CSV, TSV, etc.)

Key columns:

* `unit`: the selection label for this matrix unit.
* `sample`: the GSM accession when one can be assigned.
* `format`: the reader used for the unit.
* `preferred`: the unit selected for a sample by default.
* `has_metadata`: embedded metadata or a named sidecar was found.
* `n_files` and `files`: the grouped matrix files.

A sample can carry multiple units, such as RNA and surface protein matrices. In that case, the unit label may include a suffix such as `GSM8994520:h5ad`.

```{r formats}
table(counts[["format"]])
```

### Preferred units

Submitters often upload the same data in multiple formats. The `preferred` column marks the unit that `SeqoutMatrix()` loads when you query a GSM accession. Preference ranks the requested assay, filtered output, then format: 10x MatrixMarket, 10x HDF5, `.h5ad`, `.rds`, tar archives, and delimited tables.

Pass a `unit` label to `SeqoutMatrix()` to read another unit.

### Cell metadata detection

`has_metadata` is `TRUE` when a unit format embeds observation metadata, or when a sidecar file name looks like cell metadata or annotation. If the metadata file name contains a sample identifier, `seqout` links it to that sample. Otherwise, it is shared across the series. Metadata contents are parsed with the matrix.

## Step 2: Download and read a matrix unit

Read one matrix unit with `SeqoutMatrix()`. Downloaded files are cached under `tools::R_user_dir("seqout", "cache")` and reused across R sessions.

```{r read-rna}
rna <- SeqoutMatrix(counts, sample = "GSM8994520")
rna
```

A `seqout_matrix` has three main fields:

* `X`: counts, features by observations.
* `obs`: observation or cell annotations.
* `var`: feature or gene annotations.

In R, `X` is already oriented for Seurat and SingleCellExperiment: rows are features and columns are cells.

```{r shape}
dim(rna[["X"]])
class(rna[["X"]])
```

### Dataset classification

`kind` is inferred from row labels and shape:

* Barcode-like labels imply `single_cell`.
* GSM-like labels imply `bulk`.
* At most 32 observations implies `bulk`.
* At least 400 observations implies `single_cell`.

Datasets between 33 and 399 observations stay `unknown` because that range can include small bulk studies and Smart-seq plates.

```{r evidence}
rna[["kind"]]
rna[["evidence"]]
```

### Cell annotations

The `obs` data frame contains cell annotations deposited by the authors.

```{r obs-rna}
str(rna[["obs"]])
```

Column names come from the submitted file. In this study, `type1` is the experimental group and `type3` is the donor ID.

> [!WARNING]
> Review `var` before marker-based analysis. In this study, the author anonymized gene names.

```{r gene-names}
head(rownames(rna[["var"]]), 5)
```

## Multimodal studies

The `assay` column names the modality a unit holds, read from its file names: `rna`, `adt`, `hto` or `atac`. This study pairs gene expression with a surface antibody panel on every sample.

```{r assays}
table(counts[["assay"]], useNA = "ifany")
```

When one sample carries several modalities, `Seqout2Seurat()` reads them all and returns one object, RNA leading and the rest as extra assays on the cells they share:

```{r multimodal, message = FALSE, warning = FALSE}
mm <- Seqout2Seurat(counts, sample = "GSM8994520", max_cells = 500)

Assays(mm)
mm[["ADT"]]
```

`Seqout2SCE()` does the same, carrying the other modalities as `altExp()` entries. Pass `multimodal = FALSE` to read only the preferred unit.

A 10x multiome puts genes and peaks in one matrix and marks each row in `var$feature_type`. `Seqout2Seurat()` splits those into an `RNA` and an `ATAC` assay.

### ATAC fragments

Peak counts use one peak set. Fragments support Signac coverage tracks, re-quantification and QC. `SeqoutFragments()` lists files before download:

```{r fragments}
atac <- SeqoutListCounts("GSE156478")
SeqoutFragments(atac)[, c("sample", "file", "indexed")]
```

`download = TRUE` fetches them and fills `path`. GEO rarely deposits the `.tbi` index, so build one before handing the path to Signac:

```{r fragments-download, eval = FALSE}
frags <- SeqoutFragments(atac, sample = "GSM5065524", download = TRUE)
Rsamtools::indexTabix(frags$path[1], format = "bed")
```

To read one modality on its own, set the preference when listing:

```{r read-adt}
adt_m <- SeqoutListCounts(ACCESSION, assay = "adt") |>
  SeqoutMatrix(sample = "GSM8994520")

dim(adt_m[["X"]])
rownames(adt_m[["var"]])
```

For 10x data, `feature_type` controls the feature class filter. This ADT unit carries cell type classifications in `obs`:

```{r celltype}
head(sort(table(adt_m[["obs"]][["Cell_Type_Experimental"]]), decreasing = TRUE))
```

## Integrate study metadata

Counts and cell annotations describe the measurements. Donor demographics and study design come from `SeqoutGet()`.

```{r project}
d <- SeqoutGet(ACCESSION)

d[["meta"]][["title"]]
d[["meta"]][["overall_design"]]
```

```{r design}
design <- d[["samples"]][, c("accession", "title", "age", "Sex", "tissue")]
head(design)
```

A single matrix unit maps back to its GSM accession:

```{r join}
design[design[["accession"]] == rna[["accession"]], ]
```

Select three samples for the rest of the article. `CountsSamples()` covers the sample-to-unit join; author donor IDs live in `obs`, so the donor check uses the Seurat metadata after conversion.

```{r cross-check}
picked <- c("GSM8994520", "GSM8994523", "GSM8994533")

picked_units <- CountsSamples(counts, min_cell_count = NULL)
picked_units[picked_units[["sample"]] %in% picked, c("sample", "unit", "format")]
```

## Step 3: Create a Seurat object

`Seqout2Seurat()` accepts a parsed `seqout_matrix`, a GSM accession, a GSE accession, or the table from `SeqoutListCounts()`. Arguments in `...` go to `CreateSeuratObject()`.

```{r seurat-one, message = FALSE, warning = FALSE}
m1 <- SeqoutMatrix(counts, sample = picked[1])
Seqout2Seurat(m1, project = picked[1])
```

Cell annotations in `obs` become Seurat `meta.data`.

### Or a SingleCellExperiment

`Seqout2SCE()` creates a `SingleCellExperiment`. `X` becomes the `counts` assay, `obs` becomes `colData`, and `var` becomes `rowData`.

```{r sce-one, message = FALSE, warning = FALSE}
sce <- Seqout2SCE(m1)
sce

head(SingleCellExperiment::colData(sce), 3)
```

`SingleCellExperiment` is an optional Bioconductor dependency: `BiocManager::install("SingleCellExperiment")`. Pass `assay_name` when the matrix should land in another assay slot.

### One chain

The subject is always first, so accession-to-object calls work with the native pipe:

```{r pipe-seurat, message = FALSE, warning = FALSE}
gsm_obj <- "GSM8994520" |>
  Seqout2Seurat(project = "GSM8994520", min.cells = 3, min.features = 200)

gse_obj <- ACCESSION |>
  Seqout2Seurat(
    sample = picked[1:2], max_cells = 500,
    project = ACCESSION, min.cells = 3, min.features = 200
  )

head(gse_obj[[]][, c("sample", "type1", "type3")], 3)
```

The GSE form reads each selected unit, keeps each unit's cell metadata, binds shared features, and adds a `sample` column.

Swapping the last call for `Seqout2SCE()` returns a `SingleCellExperiment`.

`SingleCellStudies()` and the other listings that take no accession read a connection from the left. A cohort search reads its filters from the left:

```{r pipe-discover}
SeqoutConnect("api") |>
  SingleCellStudies(data = "matrix", limit = 5) |>
  subset(select = c(study_accession, study_cells, single_cell_status))

list(tissue = "blood", assay = "scRNA-seq", has_matrix = TRUE) |>
  SampleSearch(limit = 5) |>
  subset(select = c(sample, study_accession, tissue, disease))
```

### Process multiple samples

Read the selected samples in one call, cap cells per sample, then attach the study-specific covariates.

```{r seurat-build, message = FALSE, warning = FALSE}
set.seed(1)

pbmc <- Seqout2Seurat(
  counts,
  sample = picked,
  max_cells = 1500,
  sample_metadata = TRUE,
  project = ACCESSION,
  min.cells = 3,
  min.features = 200
)

pbmc[["donor"]] <- sub(".*from ([A-Z]+[0-9]+).*", "\\1", pbmc$title)
pbmc[["group"]] <- ifelse(startsWith(pbmc$donor, "MG"), "MG", "control")
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

table(pbmc$donor, pbmc$group)
```

The author donor annotations in `obs` match the study design:

```{r donor-check}
unique(pbmc[[]][, c("sample", "title", "type1", "type3")])
```

### Cohort downstream analysis

Run a standard Seurat workflow on the merged object:

```{r sctransform, message = FALSE, warning = FALSE}
pbmc <- SCTransform(pbmc, vars.to.regress = "percent.mt", verbose = FALSE)
pbmc <- RunPCA(pbmc, npcs = 30, verbose = FALSE)
pbmc <- RunUMAP(pbmc, dims = 1:30, verbose = FALSE)
pbmc <- FindNeighbors(pbmc, dims = 1:30, verbose = FALSE)
pbmc <- FindClusters(pbmc, resolution = 0.4, verbose = FALSE)

pbmc
```

Clusters:

```{r umap-clusters, fig.width = 6, fig.height = 5, message = FALSE, warning = FALSE}
DimPlot(pbmc, label = TRUE) + ggplot2::ggtitle("Clusters")
```

Study metadata:

```{r umap-metadata, fig.width = 10, fig.height = 4.5, message = FALSE, warning = FALSE}
DimPlot(pbmc, group.by = c("donor", "group", "Sex"), ncol = 3)
```

Cell composition by group:

```{r composition}
round(prop.table(table(pbmc$seurat_clusters, pbmc$group), margin = 2), 3)
```

`QuickAnnotation()` scores marker sets by cluster and assigns the highest-scoring label.

```r
markers <- list(
  t_cell = c("CD3D", "CD3E", "IL7R"),
  b_cell = c("MS4A1", "CD79A"),
  myeloid = c("LYZ", "CD14", "FCGR3A")
)

expr <- SeuratObject::LayerData(JoinLayers(pbmc), layer = "data")
labels <- QuickAnnotation(expr, pbmc$seurat_clusters, markers)
pbmc[["celltype"]] <- unname(labels[as.character(pbmc$seurat_clusters)])
```

## Batch processing and direct file access

`Matrices()` reads preferred units and returns a named list of `seqout_matrix` objects. It downloads all required files in one pass and warns for units that fail to parse.

```{r whole-study, eval = FALSE}
mats <- Matrices(counts)
```

Use `BindCounts()` when you need a matrix only. It binds shared features and returns the count matrix.

```{r bind, eval = FALSE}
merged <- BindCounts(Matrices(counts, sample = picked),
  labels = picked, max_cells = 1200
)
dim(merged)
```

Use `SeqoutCountsFiles()` to fetch local paths for custom parsers:

```{r files, eval = FALSE}
paths <- SeqoutCountsFiles(counts, sample = "GSM8994520")
```

## Bulk studies and DESeq2

Bulk series read through the same functions. A per-sample count file becomes one
observation, so `BindCounts()` returns a genes-by-samples matrix, and `CountsDesign()`
returns the study's sample characteristics as `colData`.

htseq and STAR summary rows (`__no_feature`, `N_unmapped`, and the rest) are dropped
while reading, so they never reach the library-size estimate.

```r
library(DESeq2)

counts <- SeqoutListCounts("GSE135251") # 216 bulk liver samples
cts <- BindCounts(Matrices(counts)) # 64253 genes x 216 samples
storage.mode(cts) <- "integer"

col_data <- as.data.frame(CountsDesign(counts))
rownames(col_data) <- col_data$accession
# BindCounts prefixes each column with its unit label; recover the GSM
col_data <- col_data[sub("_.*$", "", colnames(cts)), ]
rownames(col_data) <- colnames(cts)
col_data$condition <- factor(make.names(col_data[["group in paper"]]))

keep <- col_data$condition %in% c("control", "NASH_F4")
cts <- cts[, keep]
col_data <- droplevels(col_data[keep, ])
cts <- cts[rowSums(cts) >= 10, ]

dds <- DESeqDataSetFromMatrix(cts, col_data, design = ~condition)
dds <- DESeq(dds)
results(dds, contrast = c("condition", "NASH_F4", "control"))
```

## Retrieval tips

GEO supplementary data can be large.

1. Call `SeqoutListCounts()` first to inspect units and formats.
2. Pass `sample` to read selected samples or units.
3. Use `max_cells` while developing a workflow.
