---
title: "Read Counts Matrices"
description: >
  Learn how to find, download, and read counts matrices for a GEO study, join study-level metadata, and convert them to Seurat objects.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Gene Expression Omnibus (GEO) stores processed single-cell and bulk genomic datasets as supplementary files. Because GEO does not enforce a strict schema for these files, authors upload data using various formats, file names, and directory structures.

`seqout` standardizes the retrieval of these heterogeneous datasets. The package automatically identifies, groups, and parses related supplementary files into unified counts matrices.

Whether you are working with older datasets or the latest deposits, `seqout` provides a consistent three-step workflow:
1. Identify available matrices using [`SeqoutCounts()`](file:///Users/amkhrjee/Developer/source/seqout/R/R/counts.R#L51).
2. Download and read a specific matrix using [`SeqoutMatrix()`](file:///Users/amkhrjee/Developer/source/seqout/R/R/counts.R).
3. Convert the parsed matrix into a Seurat object using [`Seqout2Seurat()`](file:///Users/amkhrjee/Developer/source/seqout/R/R/counts.R).

> [!NOTE]
> Because `seqout` groups and interprets files programmatically after upload, verify the generated manifest from step 1 before proceeding to step 2.

## Prerequisites

To handle sparse matrices, `seqout` requires the `Matrix` package. To parse `.h5` and `.h5ad` formats, the package requires `hdf5r`. 

```{r install, eval = FALSE}
install.packages(c("Matrix", "hdf5r"))
```

To run the examples in this guide, load `seqout` and define a target GEO accession:

```{r setup}
library(seqout)

ACCESSION <- "GSE297547" # PBMCs from patients with myasthenia gravis and healthy controls
```

## Step 1: Identify available datasets

Query the file manifest for a given GEO study using [`SeqoutCounts()`](file:///Users/amkhrjee/Developer/source/seqout/R/R/counts.R#L51). This function queries the Seqout API to retrieve the list of files without downloading the actual dataset, which can often exceed tens of gigabytes in size.

```{r counts-table}
counts <- SeqoutCounts(ACCESSION)

nrow(counts)
head(counts[, c("unit", "sample", "format", "preferred", "has_metadata")], 8)
```

Each row in the returned table represents a **unit**, which is the smallest group of related files that `seqout` can parse as a single matrix. Common units include:
*   10x Genomics triplets (`matrix.mtx`, `barcodes.tsv`, and `features.tsv`)
*   CellRanger HDF5 files (`.h5`)
*   AnnData objects (`.h5ad`)
*   R serialized datasets (`.rds`)
*   Delimited tabular files (CSV, TSV, etc.)

### Understanding the metadata columns

The metadata table contains the following key columns:

*   `unit`: A unique identifier that represents the specific matrix unit.
*   `sample`: The GEO Sample accession (GSM), when one can be assigned.
*   `format`: The file format reader that `seqout` uses to parse the unit.
*   `preferred`: A logical flag indicating whether `seqout` selects this unit by default when querying the GSM.
*   `has_metadata`: A logical flag indicating whether cell-level or observation-level annotations are likely present.
*   `n_files` and `files`: The number of files and their respective names associated with the unit.

A single sample may contain multiple units (for example, multi-omic datasets containing both RNA and surface protein features). In these cases, the unit name includes a suffix like `GSM8994520:h5ad`.

```{r formats}
table(counts[["format"]])
```

### Preferred unit selection logic

Submitters often upload the same dataset in multiple formats, such as a CellRanger `.h5` file alongside a raw MatrixMarket triplet. The manifest includes all detected formats.

The `preferred` column marks the unit that [`SeqoutMatrix()`](file:///Users/amkhrjee/Developer/source/seqout/R/R/counts.R) loads when you query by a GSM ID instead of a specific unit label. To select a preferred unit, `seqout` ranks units for a sample based on the following criteria:

1.  **Requested Assay:** Matches the user-defined assay type.
2.  **Data Filtering:** Prioritizes filtered outputs over raw or unfiltered outputs.
3.  **File Format:** Ranks formats in the following order of preference: 10x MatrixMarket, 10x HDF5, `.h5ad`, `.rds`, compressed archives (`.tar`, `.tar.gz`), and delimited tables.

To override this default ranking and load a non-preferred unit, pass the specific `unit` identifier to `SeqoutMatrix()`.

### Cell-level metadata detection

The `has_metadata` column indicates whether cell-level metadata is likely available for a unit. This is a heuristic flag that is set to `TRUE` if:
*   The unit format is `.h5ad` or `.rds`, which natively store observation metadata.
*   The upload includes a companion file with a descriptive name suggesting cell annotations (for example, containing substrings like `metadata`, `annotation`, `celltype`, or `obs.csv`).

If the metadata file name contains a sample identifier, `seqout` links it to that specific unit. Otherwise, the package associates it with all units in the series. The metadata contents are only parsed when you load the unit.

## Step 2: Download and read a matrix unit

To download and parse a specific matrix unit, call [`SeqoutMatrix()`](file:///Users/amkhrjee/Developer/source/seqout/R/R/counts.R). The function caches downloaded files in the directory returned by `tools::R_user_dir("seqout", "cache")`. This cache persists across R sessions, so subsequent reads of the same unit do not trigger new downloads.

```{r read-rna}
rna <- SeqoutMatrix(counts, sample = "GSM8994520")
rna
```

The returned object is a `seqout_matrix` list with three primary fields:
*   `X`: The counts matrix.
*   `obs`: Observation-level (cell-level) annotations.
*   `var`: Feature-level (gene-level) annotations.

The matrix `X` follows the AnnData orientation (observations as rows, features as columns), which matches the format expected by Seurat and other single-cell analysis tools. You do not need to transpose it.

```{r shape}
dim(rna[["X"]])
class(rna[["X"]])
```

### Dataset classification: single-cell vs. bulk

`seqout` determines the type of dataset (`kind`) using the following rules:
*   If the row names match cell barcode patterns, it classifies the dataset as `single_cell`.
*   If the row names match GSM sample accessions, it classifies the dataset as `bulk`.
*   If the labels are ambiguous, the function counts the observations: datasets with 32 or fewer observations are classified as `bulk`, while datasets with 400 or more are classified as `single_cell`.

For datasets falling between 33 and 399 observations (which could represent small bulk studies or Smart-seq plates), `seqout` leaves the classification as `unknown`. You can inspect the `evidence` field for more context.

```{r evidence}
rna[["kind"]]
rna[["evidence"]]
```

### Accessing cell annotations

The `obs` data frame contains the cell-level annotations deposited by the authors. 

```{r obs-rna}
lapply(rna[["obs"]], function(x) head(unique(as.character(x)), 3))
```

Because column headers are defined by the author, they vary between datasets. In this example, the author used generic column names: `type1` (experimental group) and `type3` (donor ID).

> [!WARNING]
> Review the feature names in `var` before starting your analysis. In this specific study, the author anonymized the gene names, meaning you cannot perform marker-based cell type annotation.

```{r gene-names}
head(rownames(rna[["var"]]), 5)
```

## Load alternative assays

For multi-omic or CITE-seq studies, a single sample may contain multiple distinct assays (such as gene expression and surface antibody panels). 

By default, `SeqoutCounts()` prioritizes the RNA assay (`assay = "rna"`). You can query other assays by specifying `"adt"` (antibody-derived tags), `"hto"` (hashtag oligos), `"atac"`, or `NULL` (no preference).

`seqout` infers the assay type from file name keywords (such as `gex`, `antibody`, `hashtag`, or `peak`) and uses it to rank the available units. For 10x Genomics datasets, it filters features by their class (e.g., Gene Expression, Antibody Capture, or Peaks). To disable this filter and return all feature classes, set `feature_type = NULL`.

The following example loads the antibody-derived tag (ADT) matrix for the same sample:

```{r read-adt}
adt <- SeqoutCounts(ACCESSION, assay = "adt")
adt_m <- SeqoutMatrix(adt, sample = "GSM8994520")

dim(adt_m[["X"]])
rownames(adt_m[["var"]])
```

The parsed ADT unit includes cell type classifications under the `obs` field:

```{r celltype}
head(sort(table(adt_m[["obs"]][["Cell_Type_Experimental"]]), decreasing = TRUE))
```

## Integrate study-level metadata

The counts matrix and cell annotations represent the measurements, but they do not contain study-level information like donor demographics or the overall experimental design. To retrieve this metadata, call [`SeqoutGet()`](file:///Users/amkhrjee/Developer/source/seqout/R/R/counts.R):

```{r project}
d <- SeqoutGet(ACCESSION)

d[["meta"]][["title"]]
d[["meta"]][["overall_design"]]
```

The `samples` field contains a data frame describing the experimental design:

```{r design}
design <- d[["samples"]][, c("accession", "title", "age", "Sex", "tissue")]
head(design)
```

Because each cell in a matrix unit corresponds to a specific GSM accession, you can cross-reference the cell annotations with the study design metadata:

```{r join}
design[design[["accession"]] == rna[["accession"]], ]
```

Before proceeding, cross-check the donor IDs between the two sources:

```{r cross-check}
picked <- c("GSM8994520", "GSM8994523", "GSM8994533")

cells <- do.call(rbind, lapply(picked, function(a) {
  o <- SeqoutMatrix(counts, sample = a)[["obs"]]
  data.frame(accession = a, per_cell = unique(o[["type1"]]), donor = unique(o[["type3"]]))
}))
merge(cells, design[, c("accession", "title")], by = "accession")
```

The cell-level annotations and the study design metadata match.

## Step 3: Create a Seurat object

Use the [`Seqout2Seurat()`](file:///Users/amkhrjee/Developer/source/seqout/R/R/counts.R) function to construct a Seurat object. You can pass a parsed `seqout_matrix` object or query a single GSM accession directly. Additional arguments are forwarded to `CreateSeuratObject()`.

```{r seurat-one, message = FALSE, warning = FALSE}
library(Seurat)

Seqout2Seurat(SeqoutMatrix(counts, sample = picked[1]), project = picked[1])
```

Because the matrix `X` is already oriented correctly for Seurat, `seqout` does not transpose it. The cell annotations in `obs` are automatically imported as Seurat `meta.data`.

### Process multiple samples

To analyze multiple samples, construct individual Seurat objects, attach the sample-level covariates, and merge them:

```{r seurat-build, message = FALSE, warning = FALSE}
library(Matrix)
set.seed(1)

n_cells <- 1500

objs <- lapply(picked, function(accession) {
  m <- SeqoutMatrix(counts, sample = accession)
  keep <- sample(ncol(m[["X"]]), min(n_cells, ncol(m[["X"]])))
  m[["X"]] <- m[["X"]][, keep, drop = FALSE]
  m[["obs"]] <- m[["obs"]][keep, , drop = FALSE]
  covariates <- design[design[["accession"]] == accession, ]

  obj <- Seqout2Seurat(m, project = accession, min.cells = 3, min.features = 200)
  obj[["sample"]] <- accession
  obj[["donor"]] <- sub(".*from ([A-Z]+[0-9]+).*", "\\1", covariates[["title"]])
  obj[["group"]] <- ifelse(grepl("^MG", obj[["donor"]]), "MG", "control")
  obj[["Sex"]] <- covariates[["Sex"]]
  obj[["age"]] <- as.integer(covariates[["age"]])
  obj
})
```

Merge the individual samples and verify the cohort composition:

```{r seurat-merge, message = FALSE, warning = FALSE}
pbmc <- merge(objs[[1]], objs[-1],
  add.cell.ids = vapply(objs, function(o) unique(o[["donor"]]), character(1))
)
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

table(pbmc[["donor"]], pbmc[["group"]])
```

### Cohort downstream analysis

You can perform standard Seurat downstream workflows directly on the merged object:

```{r sctransform, message = FALSE, warning = FALSE}
pbmc <- SCTransform(pbmc, vars.to.regress = "percent.mt", verbose = FALSE)
pbmc <- RunPCA(pbmc, npcs = 30, verbose = FALSE)
pbmc <- RunUMAP(pbmc, dims = 1:30, verbose = FALSE)
pbmc <- FindNeighbors(pbmc, dims = 1:30, verbose = FALSE)
pbmc <- FindClusters(pbmc, resolution = 0.4, verbose = FALSE)

pbmc
```

Visualize the UMAP embedding colored by clusters:

```{r umap-clusters, fig.width = 6, fig.height = 5, message = FALSE, warning = FALSE}
DimPlot(pbmc, label = TRUE) + ggplot2::ggtitle("Clusters")
```

Visualize the UMAP embedding colored by the integrated study metadata:

```{r umap-metadata, fig.width = 10, fig.height = 4.5, message = FALSE, warning = FALSE}
DimPlot(pbmc, group.by = c("donor", "group", "Sex"), ncol = 3)
```

Analyze the cell composition across groups:

```{r composition}
round(prop.table(table(pbmc[["seurat_clusters"]], pbmc[["group"]]), margin = 2), 3)
```

Use the `QuickAnnotation()` helper to classify clusters based on a list of cell-type marker genes. The function scores each cell, calculates the mean score per cluster, and assigns the label with the highest score.

```{r annotation, eval = FALSE}
markers <- list(
  t_cell = c("CD3D", "CD3E", "IL7R"),
  b_cell = c("MS4A1", "CD79A"),
  myeloid = c("LYZ", "CD14", "FCGR3A")
)

expr <- SeuratObject::LayerData(JoinLayers(pbmc), layer = "data")
labels <- QuickAnnotation(expr, pbmc[["seurat_clusters"]], markers)
pbmc[["celltype"]] <- labels[as.character(pbmc[["seurat_clusters"]])]
```

## Batch processing and direct file access

For studies containing many samples, use `Matrices()` to query and load all preferred units in a single call. This downloads resources concurrently, which is faster than downloading each sample sequentially.

```{r whole-study, eval = FALSE}
mats <- Matrices(counts)
```

If a specific unit fails to load, `seqout` emits a warning and continues processing the remaining units.

To merge multiple matrices matching specific features (such as shared genes), use the `BindCounts()` helper:

```{r bind, eval = FALSE}
merged <- BindCounts(Matrices(counts, sample = picked),
  labels = picked, max_cells = 1200
)
dim(merged)
```

If you prefer to parse files using custom tools rather than loading them through `seqout`, retrieve the local file paths directly using `SeqoutCountsFiles()`:

```{r files, eval = FALSE}
paths <- SeqoutCountsFiles(counts, sample = "GSM8994520")
```

## Best practices for data retrieval

Downloading genomic datasets is resource-intensive because GEO studies can exceed tens of gigabytes. To optimize bandwidth and storage usage:

1.  **Filter First:** Call `SeqoutCounts()` first to inspect the available units and formats before downloading.
2.  **Download Selectively:** Retrieve only the specific sample accessions or units you need for your downstream analysis.
