---
title: "Download Datasets and Reads"
description: >
  Learn how to download processed files, raw sequencing reads, and alignment BAMs
  for genomic accessions, verify checksums, and manage data storage.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The `seqout` package provides three functions to download data files to disk:
*   `DownloadSupplementary()`: Downloads processed supplementary files (such as counts matrices, annotations, and archives).
*   `DownloadRuns()`: Downloads raw sequencing reads.
*   `DownloadBams()`: Downloads submitted alignment files (BAMs).

All three functions save files in a subdirectory named after the target accession. You can specify a different target directory using the `dest_dir` argument. The download functions verify checksums and automatically skip files that are already present on disk to avoid redundant downloads.

```{r setup}
library(seqout)
```

## Inspect files before downloading

Because genomic datasets can be very large (often tens of gigabytes), you should inspect the available files before initiating a download.

To list the processed supplementary files without downloading them, call `SeqoutGet()` and inspect the `supplementary` field:

```{r listing}
d <- SeqoutGet("GSE168652")

head(d$supplementary, 4)
```

Files associated directly with the series have `NA` in the `sample` column. Other files belong to specific samples.

To list the raw sequencing reads, inspect the `runs` field. The columns represent the available download locations (mirrors) provided by the archives:

```{r runs-listing}
runs <- SeqoutGet("SRP150719")$runs

nrow(runs)
head(runs[, c("run_accession", "library_layout", "fastq_bytes", "fastq_ftp")], 3)
```

For paired-end sequencing runs, both read files are packed into a single cell, separated by a semicolon. Checksums follow the same format.

## Download processed files

To download the processed files uploaded by the authors, call `DownloadSupplementary()`:

```{r supp, eval = FALSE}
DownloadSupplementary("GSE168652")
```

The scope of the download is determined by the input accession type:

| Accession Type | Download Scope |
| --- | --- |
| Series or Study (e.g., GSE, SRP) | Downloads files for the study itself and all associated samples. |
| GEO Sample (GSM) | Downloads files for that sample only. |
| Experiment, Run, or BioSample | Triggers an error. These accession types do not host processed files. |

If you pass an experiment, run, or BioSample accession, the function will stop and prompt you to query using the parent project accession instead. This prevents the tool from downloading an entire study into a directory named after a single sample or run.

```{r supp-error, error = TRUE}
DownloadSupplementary("SRR12012336")
```

Genomic archives like SRA, ENA, DDBJ, and GSA do not host processed supplementary files directly. The processed data is stored in the linked GEO or ArrayExpress records. `DownloadSupplementary()` automatically resolves these cross-archive links and downloads the files.

```{r supp-ae, eval = FALSE}
DownloadSupplementary("E-MTAB-11467") # Downloads processed archives and raw files
```

## Download sequencing reads

To download raw sequencing reads, call `DownloadRuns()`. The function resolves accessions and downloads the associated run files:
*   A run accession (SRR) downloads only that run.
*   An experiment (SRX) or sample (SRS) accession downloads all runs linked to it.
*   A study or series accession (SRP/GSE) downloads all runs for the entire study.

```{r runs, eval = FALSE}
DownloadRuns("SRR12012336")
```

### Choose a download mode

Genomic archives distribute reads in multiple formats. You can specify the format using the `mode` argument:

| Mode | Format Details |
| --- | --- |
| `"fastq"` | Standard FASTQ files hosted by ENA, verified with published MD5 checksums. |
| `"sra"` | NCBI's proprietary SRA format containing full-quality data. |
| `"sra_lite"` | NCBI's SRA format containing binned quality scores. |

If you do not specify a `mode`, the function downloads the first available format in the order listed above and prints a warning. We recommend specifying the mode explicitly:

```{r runs-mode, eval = FALSE}
DownloadRuns("SRP150719", mode = "fastq")
```

> [!WARNING]
> The `"sra_lite"` format contains binned quality scores, which leads to loss of quality detail. Use this mode only if you explicitly choose to trade quality detail for smaller file sizes.

If you pass a study accession, the function downloads every run in the study. Because this can consume substantial bandwidth, the function prints the total number of runs and estimated size before starting:

```{r runs-study, eval = FALSE}
DownloadRuns("SRP267291") # Downloads 34 runs
```

Some runs are only hosted in requester-pays cloud buckets (AWS S3 or Google Cloud Storage). Because downloading from these buckets incurs costs for the caller, specifying `mode = "s3"` or `mode = "gcs"` under anonymous access is disabled. The function will raise an error and suggest an alternative openly readable mirror when available.

### File naming and integrity

For FASTQ downloads, `DownloadRuns()` preserves the ENA naming conventions (e.g., keeping `_1` and `_2` suffixes for paired-end runs), which are expected by downstream tools.

For SRA downloads, NCBI serves files without extensions. `DownloadRuns` automatically appends the correct extension (e.g., `SRR12012336.sra` or `SRR12012336_1.sra` for multi-file runs).

If the archive publishes a checksum (which is available for all ENA FASTQ files), the function verifies the downloaded file before saving it to its final destination. If the checksum verification fails, the file is automatically deleted.

Files are downloaded using a temporary `.part` extension. This prevents incomplete or interrupted downloads from overwriting complete files.

## Download alignment files

To download the BAM alignment files uploaded by the authors, call `DownloadBams()`. These files contain the alignments mapped by the authors to their reference genome of choice and often preserve metadata (like barcode tags, methylation marks, or structural variant details) that are not present in raw reads.

To list the available BAM files before downloading, inspect the `bams` field:

```{r bams-listing}
b <- SeqoutGet("ERP117016")$bams

nrow(b)
head(b[, c("filename", "size", "md5")], 3)
```

To initiate the download, call `DownloadBams()`:

```{r bams, eval = FALSE}
DownloadBams("ERP117016")
```

Many alignment files are hosted in requester-pays cloud storage. `DownloadBams()` cannot download these files anonymously. Instead, it lists the filenames and displays the `aws` commands required to download them using your own credentials:

```{r bams-paid}
paid <- SeqoutGet("SRP071083")$bams

nrow(paid)
head(paid[, c("filename", "size", "s3_url")], 3)
```

## Archive file support matrix

The distribution of files varies across public archives:

| Archive | Processed Files | Raw Reads | Submitted Alignments |
| --- | :---: | :---: | :---: |
| GEO | Yes (Series and Samples) | Via linked SRA study | Rare |
| ArrayExpress & GEA | Yes (Processed and Raw) | Via linked ENA study | Rare |
| SRA, ENA, DDBJ | No | Yes | Sometimes |
| GSA | No | Yes | Sometimes |

`SeqoutGet()` automatically resolves the cross-references between these archives, allowing you to access all associated files using a single entry-point accession.
