---
title: "Downloads"
description: >
  Fetch the processed files, the sequencing reads and the submitted alignments
  of any accession, with checksums.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Three functions write files to disk. Each one takes any accession and resolves
it itself.

| Function | What it fetches |
| --- | --- |
| `DownloadSupplementary()` | the processed files: count matrices, annotations, archives |
| `DownloadRuns()` | the sequencing reads |
| `DownloadBams()` | the alignment files the submitter sent |

All three write into a directory named for the accession. Give `dest_dir` to
change that. All three skip a file that is already on disk, so a second call
completes the first one.

```{r setup}
library(seqout)
```

## Look before you fetch

A study can be tens of gigabytes. `SeqoutGet()` lists what exists, and it
downloads nothing.

```{r listing}
d <- SeqoutGet("GSE168652")

head(d$supplementary, 4)
```

`sample` is `NA` on the files that the series carries itself. The others belong
to one sample.

The `runs` field lists the reads. A run table carries one column per copy, and
a column is empty where the archive serves no such copy.

```{r runs-listing}
runs <- SeqoutGet("SRP150719")$runs

nrow(runs)
head(runs[, c("run_accession", "library_layout", "fastq_bytes", "fastq_ftp")], 3)
```

A paired run packs both mates into one cell, separated by a semicolon. The
checksums follow the same order.

## Processed files

`DownloadSupplementary()` fetches the files that the submitter uploaded.

```{r supp, eval = FALSE}
DownloadSupplementary("GSE168652")
```

The accession decides the scope.

| Accession | What it downloads |
| --- | --- |
| Series or study | its own files and every sample's |
| GEO sample (GSM) | that sample's files only |
| Experiment, run or BioSample | nothing; it names the project to ask for |

An experiment, a run and a BioSample carry no processed files of their own. The
function stops rather than widens to the whole study, because that would write
a study to disk under the name of one record.

```{r supp-error, error = TRUE}
DownloadSupplementary("SRR12012336")
```

SRA, ENA, DDBJ and GSA studies hold no processed files. Their files sit in the
linked GEO or ArrayExpress record, and the function follows those links for
you.

```{r supp-ae, eval = FALSE}
DownloadSupplementary("E-MTAB-11467") # processed archives and raw files
```

## Sequencing reads

`DownloadRuns()` fetches the reads. A run downloads its own files. An
experiment or a sample downloads its runs'. A series or study downloads every
run's.

```{r runs, eval = FALSE}
DownloadRuns("SRR12012336")
```

### Which copy you get

The archives publish the same reads three ways.

| `mode` | What it is |
| --- | --- |
| `"fastq"` | the ENA fastq files, with a published checksum |
| `"sra"` | NCBI's full-quality SRA copy |
| `"sra_lite"` | the same reads with binned quality scores |

Not every run is served in every form. If you name no `mode`, the function
takes the first copy each run offers, in the order above, and it warns you
which copy it took. Name `mode` to insist on one.

```{r runs-mode, eval = FALSE}
DownloadRuns("SRP150719", mode = "fastq")
```

The `"lite"` copy is last, because binned quality scores are a loss that you
should choose rather than be handed.

A study accession downloads every run of the study. That, too, is a warning
rather than an error, and it reports the number of runs and the total size
before it starts.

```{r runs-study, eval = FALSE}
DownloadRuns("SRP267291") # 34 runs
```

Some runs are served only from requester-pays buckets. curl cannot read those,
and they bill the caller, so `mode = "s3"` and `mode = "gcs"` are errors that
name the anonymous copy instead.

### Names on disk

ENA already names its files for the run, so the `_1` and `_2` suffixes of a
paired run are kept. Every downstream tool globs for them.

NCBI serves the SRA copies under names that carry no usable extension, so those
are named for their run instead: `SRR12012336.sra`, or `SRR12012336_1.sra` and
`SRR12012336_2.sra` when a run has more than one file.

### Checksums

Where the archive publishes a checksum, which it does for every fastq, the file
is verified against it before it takes its final name. A file that fails is
deleted, not kept.

Each file downloads to `<name>.part` first. An interrupted transfer therefore
never leaves a truncated file under the name of a complete one.

## Alignment files

`DownloadBams()` fetches the BAMs that the submitter sent. These are not the
same as the reads. The submitter aligned them to a reference of their own
choice, and they often carry work that the reads alone do not reconstruct:
barcode tags, methylation calls, long-read structural evidence.

```{r bams-listing}
b <- SeqoutGet("ERP117016")$bams

nrow(b)
head(b[, c("filename", "size", "md5")], 3)
```

```{r bams, eval = FALSE}
DownloadBams("ERP117016")
```

Most submitted alignments sit in requester-pays storage, which no anonymous
client can read. The function names those files rather than fetches them, and
it prints the command that would get them.

```{r bams-paid}
paid <- SeqoutGet("SRP071083")$bams

nrow(paid)
head(paid[, c("filename", "size", "s3_url")], 3)
```

Read `$bams` first when you plan a large fetch. It carries the size and the
checksum of every file, and it costs one request.

## What the archives hold

Not every archive holds every kind of file.

| Archive | Processed files | Reads | Alignments |
| --- | :---: | :---: | :---: |
| GEO | yes, series and sample | through the linked SRA study | rare |
| ArrayExpress and GEA | yes, processed and raw | through the linked ENA study | rare |
| SRA, ENA, DDBJ | no | yes | sometimes |
| GSA | no | yes | sometimes |

`SeqoutGet()` follows the links between them, so one accession is enough
whichever archive you start from.
