---
title: "Download Datasets and Reads"
description: >
  Learn how to download processed files, raw sequencing reads, and alignment BAMs
  for genomic accessions, verify checksums, and manage data storage.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Three functions download files to disk:

*   `DownloadSupplementary()`: processed supplementary files (counts matrices, annotations, archives).
*   `DownloadRuns()`: raw sequencing reads.
*   `DownloadBAMs()`: submitted alignment files (BAMs).

Each saves into a subdirectory named after the accession unless `dest_dir` names another. All three verify checksums and skip files already on disk.

```{r setup}
library(seqout)
```

## Inspect files before downloading

Datasets often run to tens of gigabytes, so inspect the files first. The `supplementary` field of `SeqoutGet()` lists processed supplementary files without downloading them:

```{r listing}
d <- SeqoutGet("GSE168652")

head(d$supplementary, 4)
```

Files associated directly with the series have `NA` in the `sample` column. Other files belong to specific samples.

For GEO, ArrayExpress and DDBJ GEA projects, `SupplementaryFiles()` lists the same files with a ready-made download command and the result of seqout's scan for Excel gene-name damage (`"SEPT1"` turned into `"1-Sep"`). A `NULL` `gene_corruption` means the file has not been scanned yet, not that it is clean:

```{r supp-files}
sf <- SupplementaryFiles("GSE168652")
sf[, c("filename", "download_command")]
```

The `runs` field lists raw sequencing reads, including a URL column for each archive mirror:

```{r runs-listing}
runs <- SeqoutGet("SRP150719")$runs

nrow(runs)
head(runs[, c("run_accession", "library_layout", "fastq_bytes", "fastq_ftp")], 3)
```

Paired-end runs pack both read files into one cell, separated by a semicolon. Checksums follow the same format.

`RunFiles()` unpacks those cells into one row per file, with its size and checksum. It picks the same copy `DownloadRuns()` would fetch, so you can total a download before you start it:

```{r run-files}
files <- RunFiles(runs, mode = "fastq")
head(files, 4)
sprintf("%d files, %.1f GB", nrow(files), sum(files$bytes, na.rm = TRUE) / 1e9)
```

## Download processed files

`DownloadSupplementary()` fetches the processed files uploaded by the authors:

```{r supp, eval = FALSE}
DownloadSupplementary("GSE168652", "GSE168652")
```

The accession type sets the scope:

| Accession Type | Download Scope |
| --- | --- |
| Series or Study (e.g., GSE, SRP) | The study and all its samples. |
| GEO Sample (GSM) | That sample only. |
| Experiment, Run, or BioSample | An error. These accessions host no processed files. |

The error asks for the parent project accession, so a whole study never lands in a directory named after one sample or run.

```{r supp-error, error = TRUE}
DownloadSupplementary("SRR12012336", "SRR12012336")
```

SRA, ENA, DDBJ, and GSA host no processed supplementary files; those live in the linked GEO or ArrayExpress records. `DownloadSupplementary()` follows the link and downloads them.

```{r supp-ae, eval = FALSE}
DownloadSupplementary("E-MTAB-11467", "E-MTAB-11467") # Downloads processed archives and raw files
```

## Download sequencing reads

`DownloadRuns()` downloads raw sequencing reads. The accession sets the scope:

*   A run accession (SRR) downloads only that run.
*   An experiment (SRX) or sample (SRS) accession downloads all runs linked to it.
*   A study or series accession (SRP/GSE) downloads all runs for the entire study.

```{r runs, eval = FALSE}
DownloadRuns("SRR12012336", "SRR12012336")
```

### Choose a download mode

Archives distribute reads in several formats; `mode` picks one:

| Mode | Format Details |
| --- | --- |
| `"fastq"` | FASTQ files hosted by ENA, verified against published MD5 checksums. |
| `"sra"` | NCBI's SRA format with full-quality data. |
| `"sra_lite"` | NCBI's SRA format with binned quality scores. |

Without `mode`, the function downloads the first available format in the order above and prints a warning. Set it explicitly:

```{r runs-mode, eval = FALSE}
DownloadRuns("SRP150719", "SRP150719", mode = "fastq")
```

> [!WARNING]
> Binned quality scores lose detail. Use `"sra_lite"` only to trade that detail for smaller files.

For a study accession, the function prints the run count and estimated size before starting:

```{r runs-study, eval = FALSE}
DownloadRuns("SRP267291", "SRP267291") # Downloads 34 runs
```

Some runs are hosted only in requester-pays cloud buckets (AWS S3 or Google Cloud Storage), which bill the caller. `mode = "s3"` and `mode = "gcs"` are refused under anonymous access; the function raises an error and suggests an openly readable mirror when one exists.

### File naming and integrity

FASTQ downloads keep ENA's file names (e.g., the `_1` and `_2` suffixes of paired-end runs), which downstream tools expect.

NCBI serves SRA files without extensions; `DownloadRuns` appends one (e.g., `SRR12012336.sra` or `SRR12012336_1.sra` for multi-file runs).

When the archive publishes a checksum (all ENA FASTQ files have one), the function verifies the file before moving it to its final destination and deletes it if verification fails.

Downloads write to a temporary `.part` file, so an interrupted download never overwrites a complete one.

## Download alignment files

`DownloadBAMs()` fetches the BAM files uploaded by the authors. These hold the authors' alignments to their chosen reference and often keep metadata absent from raw reads, such as barcode tags, methylation marks, or structural variant details.

The `bams` field lists them before download:

```{r bams-listing}
b <- SeqoutGet("ERP117016")$bams

nrow(b)
head(b[, c("filename", "size", "md5")], 3)
```

```{r bams, eval = FALSE}
DownloadBAMs("ERP117016", "ERP117016")
```

Many alignment files sit in requester-pays cloud storage, which `DownloadBAMs()` cannot fetch anonymously. For those it lists the filenames and prints the `aws` commands to download them with your own credentials:

```{r bams-paid}
paid <- SeqoutGet("SRP071083")$bams

nrow(paid)
head(paid[, c("filename", "size", "s3_url")], 3)
```

## Download the database dump

The Parquet backend reads the published dump remotely. For fully offline work, `DownloadDump()` fetches the tables to disk; it reports the total size first. `Tables()` names what you can ask for:

```{r dump, eval = FALSE}
DownloadDump(dest_dir = "~/seqout-dump", tables = c("unified_metadata", "geo_series"))

con <- SeqoutConnect("parquet", data_dir = "~/seqout-dump")
Query("SELECT count(*) FROM unified_metadata", con = con)
```

## Archive file support matrix

| Archive | Processed Files | Raw Reads | Submitted Alignments |
| --- | :---: | :---: | :---: |
| GEO | Yes (Series and Samples) | Via linked SRA study | Rare |
| ArrayExpress & GEA | Yes (Processed and Raw) | Via linked ENA study | Rare |
| SRA, ENA, DDBJ | No | Yes | Sometimes |
| GSA | No | Yes | Sometimes |

`SeqoutGet()` resolves the cross-references between these archives, so one accession reaches all associated files.
