---
title: "Retrieve Dataset Metadata"
description: >
  Query and explore study records, samples, runs, harmonized attributes, 
  publications, and read-derived quality metrics.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`SeqoutGet()` resolves an archive accession to dataset metadata and related records across archives.

```{r setup}
library(seqout)

d <- SeqoutGet("GSE168652")
d
```

Fields load on first access and cache their results for later reads.

The dataset object contains the following fields:

```{r fields}
names(d)
```

| Field | Contents |
| --- | --- |
| `meta` | Study-level metadata: title, abstract, design description, submission dates, and files. |
| `samples` | A data frame containing one row per sample in the study. |
| `experiments` | A data frame containing library preparation details for each experiment. |
| `runs` | A data frame containing sequencing runs and download URLs. |
| `supplementary` | A list of processed supplementary files associated with the study and its samples. |
| `bams` | Alignment files submitted by the authors. |
| `pubs` | Publications linked to the dataset. |
| `links` | Accession mappings for the same dataset in other archives. |
| `enriched` | Harmonized sample attributes generated by `seqout.org`. |
| `detail` | The specific record detail if `SeqoutGet()` was called with a sample or run accession. |
| `kind`, `project`, `geo`, `sra` | Source metadata classification and mapped parent accessions. |

## Study-level metadata

List study metadata fields with `names()`:

```{r meta-names}
names(d$meta)
```

You can access these fields directly:

```{r meta}
d$meta$title
d$meta$overall_design
c(published = d$meta$published_at, organisms = d$meta$organisms)
```

## Samples, experiments, and runs

`samples` has one row per sample, with submitter attributes unpacked into columns:

```{r samples}
d$samples
```

The `experiments` field contains library preparation metadata, and the `runs` field contains sequencing run metadata along with target file URLs:

```{r experiments}
d$experiments[, c("accession", "library_strategy", "library_layout", "instrument_model")]
```

```{r runs}
d$runs[, c("run_accession", "experiment_accession", "ncbi_sra_url_aws")]
```

Because public archives do not distribute all runs in every format, some columns may contain `NA`. For example, this study does not host ENA FASTQ files, so the `fastq_ftp` column is empty:

```{r runs-fastq}
d$runs[, c("run_accession", "fastq_ftp", "ncbi_sra_normalized_bytes")]
```

To let the package select the optimal format for downloads, see [Downloads](downloads.html).

## Cross-archive accession mapping

Authors often deposit the same study in multiple archives. The `links` field lists all resolved accessions for the dataset:

```{r links}
d$links
```

You can pass any of these accessions to `SeqoutGet()` to retrieve the same dataset object. 

To identify the type of accession without making an API request, call `AccessionKind()`:

```{r kind}
AccessionKind("GSE168652")
AccessionKind("SRR12012336")
AccessionKind("E-MTAB-11467")
```

## Harmonized sample metadata

`seqout.org` maps free-text sample attributes to a shared vocabulary of tissues, diseases, cell types, assays, and developmental stages.

You can access these standardized attributes in the `enriched` field:

```{r enriched}
d$enriched[, c("sample", "tissue", "disease", "assay", "assay_category")]
```

Each harmonized field includes a corresponding ontology identifier (CURIE) from major biological ontologies (such as UBERON, MONDO, and CL), allowing you to group and compare samples across studies:

```{r enriched-ids}
d$enriched[, c("sample", "disease_ontology_id", "disease_ontology_name")]
```

To build cohorts across studies using these standardized fields, see [Search](search.html).

## Mapped publications

The `pubs` field lists the publications linked to the dataset, including journal names and citation counts:

```{r pubs}
d$pubs[, c("pmid", "title", "journal", "citation_count")]
```

[Search](search.html) covers publication and author lookups with `Paper()` and `Author()`.

## Retrieve summaries for multiple accessions

To fetch abbreviated records for multiple projects in a single API call, use `Summaries()`:

```{r summaries}
Summaries(c("GSE168652", "GSE151530", "SRP267291"))
```

## Read-derived quality metrics

In addition to parsing submitter-declared metadata, `seqout` screens the raw sequencing reads to verify metadata and assess sample composition.

### Single-cell metadata

Call `ProjectSingleCell()` to retrieve read-derived single-cell metrics, such as cell and gene counts, predicted assay categories, and sex verification verdicts:

```{r single-cell}
sc <- ProjectSingleCell("GSE168652")
sc[, c("sample_accession", "cells", "genes", "assay_display", "sex_verdict")]
```

> [!NOTE]
> The `cells` field counts columns in the raw expression matrix. For unfiltered 10x Genomics datasets, this is all detected barcodes and an upper bound for final cells.

### Microbial screening

The package also screens reads for microbial sequences. The `has_viral_reads` and `has_bacterial_reads` fields indicate whether the screen detected microbial content passing quality threshold gates. A value of `NA` indicates that the sample has not been screened.

```{r flags}
sc[, c("sample_accession", "has_viral_reads", "hpv_top_type", "n_flags")]
```

To inspect all microbial detections (including those below the quality gate threshold), call `SampleMicrobes()`:

```{r microbes}
m <- SampleMicrobes("GSM5155196", kind = "viral")
m[, c("organism", "n_unitigs", "max_breadth_frac")]
```

`max_breadth_frac` is the fraction of the reference genome covered. Detections below the reporting gate leave `has_viral_reads` as `FALSE`.

To filter out low-confidence detections and keep only those that pass the quality gate, set `validated_only = TRUE`. The function automatically excludes common spike-ins, negative controls, and background contaminants unless you set `include_background = TRUE`.

## Retrieve sample or run details

If you query `SeqoutGet()` using a sample (GSM) or run (SRR) accession, the returned dataset object includes the specific record details in the `detail` field:

```{r detail}
s <- SeqoutGet("GSM5155196")
s$kind
s$detail[, c("accession", "title", "tissue", "platform_ref")]
```

The other fields resolve to the parent study:

```{r detail-project}
s$project
nrow(s$samples)
```
