---
title: "Metadata"
description: >
  One accession, every record below it: the study, the samples, the runs, the
  ontology terms, the papers and the read-derived calls.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`SeqoutGet()` is the one entry point for metadata. It takes any accession and
resolves the rest itself.

```{r setup}
library(seqout)

d <- SeqoutGet("GSE168652")
d
```

Each field makes its request the first time you read it, and it keeps the
answer. A second read of the same field costs nothing. A field that you never
touch costs nothing.

```{r fields}
names(d)
```

| Field | What it holds |
| --- | --- |
| `meta` | the project record: title, summary, design, dates, files |
| `samples` | one row per sample |
| `experiments` | one row per library preparation |
| `runs` | one row per sequencing run, with the file URLs |
| `supplementary` | one row per supplementary file, for the series and the samples |
| `bams` | the alignment files the submitter sent |
| `pubs` | the publications linked to the dataset |
| `links` | the same dataset in the other archives |
| `enriched` | the ontology terms for each sample |
| `detail` | the record of the accession itself, for a sample or a run |
| `kind`, `project`, `geo`, `sra` | where the data sits |

## The study record

```{r meta-names}
names(d$meta)
```

A project record is too wide to print. Read the fields you want by name.

```{r meta}
d$meta$title
d$meta$overall_design
c(published = d$meta$published_at, organisms = d$meta$organisms)
```

## Samples, experiments and runs

`samples` holds one row per sample. The sample attributes are unnested, so a
submitter's own keys become columns.

```{r samples}
d$samples
```

The same field answers for SRA, ENA, DDBJ and GSA. `experiments` holds the
library preparations, and `runs` holds the sequencing runs with their file
URLs.

```{r experiments}
d$experiments[, c("accession", "library_strategy", "library_layout", "instrument_model")]
```

```{r runs}
d$runs[, c("run_accession", "experiment_accession", "ncbi_sra_url_aws")]
```

A run table carries one column per copy of the reads. Not every run is served
in every form, so a column can be empty. This study has no ENA fastq copy, and
`fastq_ftp` is therefore `NA` for both runs.

```{r runs-fastq}
d$runs[, c("run_accession", "fastq_ftp", "ncbi_sra_normalized_bytes")]
```

[Downloads](downloads.html) chooses between these copies for you.

## The same dataset in other archives

A study is deposited more than once. `links` names the records that hold the
same data.

```{r links}
d$links
```

`SeqoutGet()` accepts any of those accessions and returns the same dataset.
`AccessionKind()` says what an accession is, without a request.

```{r kind}
AccessionKind("GSE168652")
AccessionKind("SRR12012336")
AccessionKind("E-MTAB-11467")
```

## Ontology terms

seqout reads the free text of each sample and adds ontology terms to it. The
terms give the tissue, the disease, the cell type, the assay and the
development stage.

```{r enriched}
d$enriched[, c("sample", "tissue", "disease", "assay", "assay_category")]
```

Each term carries its ontology ID.

```{r enriched-ids}
d$enriched[, c("sample", "disease_ontology_id", "disease_ontology_name")]
```

[Search](search.html) searches these terms across every study.

## Publications

`pubs` holds the papers that name the dataset, with the journal and the
citation count.

```{r pubs}
d$pubs[, c("pmid", "title", "journal", "citation_count")]
```

`Paper()` is the reverse lookup. It takes a PubMed ID or a DOI and returns the
datasets that name that paper.

```{r paper}
Paper(pmid = "34764296")
```

`Author()` does the same for a name.

```{r author}
head(Author("Aviv Regev", limit = 5)[, c("accession", "title", "journal")], 5)
```

## Many accessions at once

`Summaries()` reads short records for many projects in one request.

```{r summaries}
Summaries(c("GSE168652", "GSE151530", "SRP267291"))
```

## What the reads contain

The fields above report what the submitter declared. seqout also screens the
reads themselves. `ProjectSingleCell()` returns the matrix dimensions and the
read-derived calls for each sample of a study.

```{r single-cell}
sc <- ProjectSingleCell("GSE168652")
sc[, c("sample_accession", "cells", "genes", "assay_display", "sex_verdict")]
```

`cells` counts the columns of the matrix. For an unfiltered 10x matrix these
are barcodes, so the number is an upper bound and a sum overcounts.

`has_viral_reads` and `has_bacterial_reads` are `NA` when the sample was never
screened, and `FALSE` when the screen found no gated hit.

```{r flags}
sc[, c("sample_accession", "has_viral_reads", "hpv_top_type", "n_flags")]
```

`SampleMicrobes()` shows the detections behind those flags, one row per
organism. It shows every detection, and not only the ones that pass the gate,
so a sample whose flag is `FALSE` can still list organisms here.

```{r microbes}
m <- SampleMicrobes("GSM5155196", kind = "viral")
m[, c("organism", "n_unitigs", "max_breadth_frac")]
```

`max_breadth_frac` is the fraction of the reference that the reads cover. The
HPV16 detection above covers 17% of it, which is below the gate, so
`has_viral_reads` stays `FALSE`. Set `validated_only = TRUE` to keep only the
detections that pass.

The spike-in control and the negative control are never summed into the totals.
Reagent and skin organisms are excluded, unless you set
`include_background = TRUE`.

## One sample or one run

`SeqoutGet()` also takes a sample accession or a run accession. The `detail`
field then holds the record of that accession itself.

```{r detail}
s <- SeqoutGet("GSM5155196")
s$kind
s$detail[, c("accession", "title", "tissue", "platform_ref")]
```

The other fields still answer for the study that the sample belongs to.

```{r detail-project}
s$project
nrow(s$samples)
```
