---
title: "Retrieve Dataset Metadata"
description: >
  Query and explore study records, samples, runs, harmonized attributes, 
  publications, and read-derived quality metrics.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The `SeqoutGet()` function serves as the single entry point for retrieving dataset metadata. It queries the `seqout.org` API using any supported database accession and retrieves all associated records across public archives.

```{r setup}
library(seqout)

d <- SeqoutGet("GSE168652")
d
```

To conserve system resources and network bandwidth, the dataset object loads fields lazily. The package fetches data from the API only when you access a field for the first time, and caches the result for subsequent reads. Fields you do not access incur no API requests.

The dataset object contains the following fields:

```{r fields}
names(d)
```

| Field | Contents |
| --- | --- |
| `meta` | Study-level metadata: title, abstract, design description, submission dates, and files. |
| `samples` | A data frame containing one row per sample in the study. |
| `experiments` | A data frame containing library preparation details for each experiment. |
| `runs` | A data frame containing sequencing runs and download URLs. |
| `supplementary` | A list of processed supplementary files associated with the study and its samples. |
| `bams` | Alignment files submitted by the authors. |
| `pubs` | Publications linked to the dataset. |
| `links` | Accession mappings for the same dataset in other archives. |
| `enriched` | Harmonized sample attributes generated by `seqout.org`. |
| `detail` | The specific record detail if `SeqoutGet()` was called with a sample or run accession. |
| `kind`, `project`, `geo`, `sra` | Source metadata classification and mapped parent accessions. |

## Study-level metadata

To inspect the attributes available in the study-level metadata, check the names of the `meta` field:

```{r meta-names}
names(d$meta)
```

You can access these fields directly:

```{r meta}
d$meta$title
d$meta$overall_design
c(published = d$meta$published_at, organisms = d$meta$organisms)
```

## Samples, experiments, and runs

The `samples` field contains a data frame where each row represents a single sample. The package automatically unpacks author-defined attributes into flat columns:

```{r samples}
d$samples
```

The `experiments` field contains library preparation metadata, and the `runs` field contains sequencing run metadata along with target file URLs:

```{r experiments}
d$experiments[, c("accession", "library_strategy", "library_layout", "instrument_model")]
```

```{r runs}
d$runs[, c("run_accession", "experiment_accession", "ncbi_sra_url_aws")]
```

Because public archives do not distribute all runs in every format, some columns may contain `NA`. For example, this study does not host ENA FASTQ files, so the `fastq_ftp` column is empty:

```{r runs-fastq}
d$runs[, c("run_accession", "fastq_ftp", "ncbi_sra_normalized_bytes")]
```

To let the package select the optimal format for downloads, see [Downloads](downloads.html).

## Cross-archive accession mapping

Authors often deposit the same study in multiple archives. The `links` field lists all resolved accessions for the dataset:

```{r links}
d$links
```

You can pass any of these accessions to `SeqoutGet()` to retrieve the same dataset object. 

To identify the type of accession without making an API request, call `AccessionKind()`:

```{r kind}
AccessionKind("GSE168652")
AccessionKind("SRR12012336")
AccessionKind("E-MTAB-11467")
```

## Harmonized sample metadata

Because authors describe sample attributes using free-text fields, metadata varies widely across studies. `seqout.org` parses these text attributes and maps them to a standardized vocabulary covering tissue types, diseases, cell types, assay methods, and developmental stages.

You can access these standardized attributes in the `enriched` field:

```{r enriched}
d$enriched[, c("sample", "tissue", "disease", "assay", "assay_category")]
```

Each harmonized field includes a corresponding ontology identifier (CURIE) from major biological ontologies (such as UBERON, MONDO, and CL), allowing you to group and compare samples across studies:

```{r enriched-ids}
d$enriched[, c("sample", "disease_ontology_id", "disease_ontology_name")]
```

To build cohorts across studies using these standardized fields, see [Search](search.html).

## Mapped publications

The `pubs` field lists the publications linked to the dataset, including journal names and citation counts:

```{r pubs}
d$pubs[, c("pmid", "title", "journal", "citation_count")]
```

This field maps a dataset to its publications. To perform the reverse query (mapping a publication or author to their datasets), use the `Paper()` and `Author()` functions documented in [Search](search.html).

## Retrieve summaries for multiple accessions

To fetch abbreviated records for multiple projects in a single API call, use `Summaries()`:

```{r summaries}
Summaries(c("GSE168652", "GSE151530", "SRP267291"))
```

## Read-derived quality metrics

In addition to parsing submitter-declared metadata, `seqout` screens the raw sequencing reads to verify metadata and assess sample composition.

### Single-cell metadata

Call `ProjectSingleCell()` to retrieve read-derived single-cell metrics, such as cell and gene counts, predicted assay categories, and sex verification verdicts:

```{r single-cell}
sc <- ProjectSingleCell("GSE168652")
sc[, c("sample_accession", "cells", "genes", "assay_display", "sex_verdict")]
```

> [!NOTE]
> The `cells` field counts the total number of columns in the raw expression matrix. For unfiltered 10x Genomics datasets, this represents all detected barcodes and serves as an upper bound rather than the final cell count.

### Microbial screening

The package also screens reads for microbial sequences. The `has_viral_reads` and `has_bacterial_reads` fields indicate whether the screen detected microbial content passing quality threshold gates. A value of `NA` indicates that the sample has not been screened.

```{r flags}
sc[, c("sample_accession", "has_viral_reads", "hpv_top_type", "n_flags")]
```

To inspect all microbial detections (including those below the quality gate threshold), call `SampleMicrobes()`:

```{r microbes}
m <- SampleMicrobes("GSM5155196", kind = "viral")
m[, c("organism", "n_unitigs", "max_breadth_frac")]
```

The `max_breadth_frac` column indicates the fraction of the reference genome covered by the reads. In this example, the HPV16 detection covers 17.3% of the genome, which is below the reporting gate, meaning `has_viral_reads` remains `FALSE`.

To filter out low-confidence detections and keep only those that pass the quality gate, set `validated_only = TRUE`. The function automatically excludes common spike-ins, negative controls, and background contaminants unless you set `include_background = TRUE`.

## Retrieve sample or run details

If you query `SeqoutGet()` using a sample (GSM) or run (SRR) accession, the returned dataset object includes the specific record details in the `detail` field:

```{r detail}
s <- SeqoutGet("GSM5155196")
s$kind
s$detail[, c("accession", "title", "tissue", "platform_ref")]
```

The other fields in the dataset object still resolve to the parent study to provide context:

```{r detail-project}
s$project
nrow(s$samples)
```
