---
title: "Retrieve Dataset Metadata"
description: >
  Query and explore study records, samples, runs, harmonized attributes, 
  publications, and read-derived quality metrics.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`SeqoutGet()` resolves an archive accession to dataset metadata and related records across archives.

```{r setup}
library(seqout)

d <- SeqoutGet("GSE168652")
d
```

Fields load on first access and are cached for later reads:

```{r fields}
names(d)
```

| Field | Contents |
| --- | --- |
| `meta` | Study-level metadata: title, abstract, design description, submission dates, and files. |
| `samples` | One row per sample. |
| `experiments` | Library preparation details, one row per experiment. |
| `runs` | Sequencing runs and download URLs. |
| `supplementary` | Processed supplementary files for the study and its samples. |
| `bams` | Alignment files submitted by the authors. |
| `pubs` | Publications linked to the dataset. |
| `links` | Accession mappings for the same dataset in other archives. |
| `enriched` | Harmonized sample attributes generated by `seqout.org`. |
| `detail` | The record itself when `SeqoutGet()` was called with a sample or run accession. |
| `kind`, `project`, `geo`, `sra` | Source metadata classification and mapped parent accessions. |

## Study-level metadata

```{r meta-names}
names(d$meta)
```

```{r meta}
d$meta$title
d$meta$overall_design
c(published = d$meta$published_at, organisms = d$meta$organisms)
```

## Samples, experiments, and runs

`samples` has one row per sample, with submitter attributes unpacked into columns:

```{r samples}
d$samples
```

```{r experiments}
d$experiments[, c("accession", "library_strategy", "library_layout", "instrument_model")]
```

```{r runs}
d$runs[, c("run_accession", "experiment_accession", "ncbi_sra_url_aws")]
```

Archives do not distribute every run in every format, so some columns may be `NA`. This study has no ENA FASTQ files, so `fastq_ftp` is empty:

```{r runs-fastq}
d$runs[, c("run_accession", "fastq_ftp", "ncbi_sra_normalized_bytes")]
```

[Downloads](downloads.html) covers choosing a format.

## Cross-archive accession mapping

Authors often deposit the same study in several archives. `links` lists every resolved accession:

```{r links}
d$links
```

`SeqoutGet()` returns the same dataset for any of them.

For one specific link, four functions each follow a single hop and return one accession (`NA` when nothing links):

```{r resolve}
ResolveStudy("SRR13927092") # run, experiment or sample -> its study
GSMSeries("GSM5155196") # GEO sample -> its series
LinkedStudy("GSE168652") # series -> the sequencing study holding its runs
LinkedGEO("SRP310948") # sequencing study -> the series holding its processed files
```

`AccessionKind()` classifies an accession without an API request:

```{r kind}
AccessionKind("GSE168652")
AccessionKind("SRR12012336")
AccessionKind("E-MTAB-11467")
```

## Harmonized sample metadata

`seqout.org` maps free-text sample attributes to a shared vocabulary of tissues, diseases, cell types, assays, and developmental stages.

They are in the `enriched` field:

```{r enriched}
d$enriched[, c("sample", "tissue", "disease", "assay", "assay_category")]
```

Each harmonized field carries an ontology identifier (CURIE) from ontologies such as UBERON, MONDO, and CL, so you can group and compare samples across studies:

```{r enriched-ids}
d$enriched[, c("sample", "disease_ontology_id", "disease_ontology_name")]
```

To build cohorts across studies using these standardized fields, see [Search](search.html).

## Mapped publications

`pubs` lists linked publications with journal names and citation counts:

```{r pubs}
d$pubs[, c("pmid", "title", "journal", "citation_count")]
```

[Search](search.html) covers publication and author lookups with `Paper()` and `Author()`.

## Retrieve summaries for multiple accessions

`Summaries()` fetches abbreviated records for several projects in one API call:

```{r summaries}
Summaries(c("GSE168652", "GSE151530", "SRP267291"))
```

## Read-derived quality metrics

Beyond submitter-declared metadata, `seqout` screens the raw reads to verify metadata and assess sample composition.

### Single-cell metadata

`ProjectSingleCell()` returns read-derived single-cell metrics such as cell and gene counts, predicted assay categories, and sex verification verdicts:

```{r single-cell}
sc <- ProjectSingleCell("GSE168652")
sc[, c("sample_accession", "cells", "genes", "assay_display", "sex_verdict")]
```

> [!NOTE]
> The `cells` field counts columns in the raw expression matrix. For unfiltered 10x Genomics datasets, this is all detected barcodes and an upper bound for final cells.

To ask only whether a study is single-cell, and on what evidence, call `ProjectSingleCellStatus()`. `ProjectSingleCellSummary()` rolls the per-sample metrics up to one study-level row:

```{r single-cell-study}
status <- ProjectSingleCellStatus("GSE168652")
status[, c("accession", "single_cell_status", "n_independent_evidence", "has_matrix")]

s <- ProjectSingleCellSummary("GSE168652")
s[, c("study_accession", "study_cells", "n_samples_reported", "any_unfiltered")]
```

### Long-read chemistry

For PacBio and Oxford Nanopore studies, `ProjectLongreadChemistry()` lists every long-read run with its instrument and sequencing chemistry:

```{r longread-chemistry}
LongreadProjects(technology = "PacBio", sort = "n_runs", order = "asc", limit = 1)$study_accession |>
  ProjectLongreadChemistry()
```

### Microbial screening

Reads are also screened for microbial sequences. `has_viral_reads` and `has_bacterial_reads` flag microbial content that passed the screen's quality gates; `NA` means the sample has not been screened.

```{r flags}
sc[, c("sample_accession", "has_viral_reads", "hpv_top_type", "n_flags")]
```

`SampleMicrobes()` lists every microbial detection, including those below the quality gate:

```{r microbes}
m <- SampleMicrobes("GSM5155196", kind = "viral")
m[, c("organism", "n_unitigs", "max_breadth_frac")]
```

`max_breadth_frac` is the fraction of the reference genome covered. Detections below the reporting gate leave `has_viral_reads` as `FALSE`.

`validated_only = TRUE` keeps only detections that pass the quality gate. Common spike-ins, negative controls, and background contaminants are excluded unless you set `include_background = TRUE`.

## Retrieve sample or run details

Called with a sample (GSM) or run (SRR) accession, `SeqoutGet()` puts that record in `detail`:

```{r detail}
s <- SeqoutGet("GSM5155196")
s$kind
s$detail[, c("accession", "title", "tissue", "platform_ref")]
```

The other fields resolve to the parent study:

```{r detail-project}
s$project
nrow(s$samples)
```
