---
title: "Search Studies and Samples"
description: >
  Learn how to query genomic datasets, apply metadata filters, perform structured 
  boolean searches, resolve ontology terms, and construct sample cohorts.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

The `seqout` package provides four functions to search and filter database records:
*   `SeqoutSearch()`: Searches for studies matching text queries or filters.
*   `SampleSearch()`: Searches for individual samples to compile cohorts.
*   `Paper()`: Locates datasets linked to a specific publication.
*   `Author()`: Locates datasets linked to a specific researcher.

All search functions read data from the REST API.

```{r setup}
library(seqout)
```

## Search for studies

To search for studies across all seven genomic archives, call `SeqoutSearch()`. The function matches your search terms against the title, abstract, and experimental design description of each study.

```{r first-search}
SeqoutSearch("liver cancer scRNA", limit = 5)
```

The `source` column indicates which archive hosts each study.

### Apply search filters

You can narrow your search by specifying filter arguments. You do not need to supply a query string when using filters.

```{r filters-only}
SeqoutSearch(organism = "Mus musculus", assay_l1 = "Transcriptomic", limit = 5)
```

You can combine a search query with filters:

```{r filters}
SeqoutSearch(
  "liver",
  organism         = "Homo sapiens",
  library_strategy = "RNA-Seq",
  limit            = 5
)
```

### Supported search filters

You can combine any of the following filters in a query:

| Filter | Selection Criteria |
| --- | --- |
| `db` or `source` | Limit results to a specific archive (e.g., `geo`, `sra`). |
| `organism` | Scientific name of the target organism. |
| `library_strategy`, `library_source`, `platform`, `instrument_model` | Methodological details of library preparation and sequencing. |
| `assay_l1`, `assay_l2` | High-level assay category and specific protocol method. |
| `country`, `geo_country_code_iso2`, `geo_lat`, `geo_lng`, `geo_radius_km` | Geographic origin of the study. |
| `journal` | Publishing journal of the linked publication. |
| `multi_platform` | Mapped studies that use more than one sequencing platform. |
| `date_from`, `date_to` | Date range of the last database modification (`yyyy-mm-dd`). |

> [!WARNING]
> You cannot combine the `library_source` filter with `assay_l1`, `assay_l2`, or any geographic (`geo_*`) filters because the underlying search indexes are incompatible.

If you specify `db` or `source` on its own, you must supply at least one other search term or filter:

```{r one-archive}
SeqoutSearch("hepatocellular carcinoma", db = "geo", limit = 5)
```

The `assay_l1` filter represents broad strategies (e.g., `"Transcriptomic"`), whereas `assay_l2` represents specific assay protocols (e.g., `"ATAC-seq"`):

```{r assay}
SeqoutSearch("liver", assay_l2 = "ATAC-seq", limit = 5)
```

### Date filtering

To retrieve studies updated within a specific timeframe, use the `date_from` and `date_to` filters:

```{r dates}
SeqoutSearch("liver cancer",
  date_from = "2024-01-01", date_to = "2024-01-31", limit = 5
)
```

```{r dates-combined}
nrow(SeqoutSearch("liver", assay_l1 = "Transcriptomic", date_from = "2025-01-01"))
```

## Structured boolean searches

To construct complex queries, write boolean search expressions. You can group terms using parentheses `()`, search for exact phrases using double quotes `""`, use `*` for prefix wildcard matching, and join terms using the uppercase operators `AND`, `OR`, and `NOT`.

```{r boolean}
SeqoutSearch('("aging" OR "aged") (gut OR colon) immun*', limit = 5)
```

By default, adjacent terms are joined with an implicit `AND` operator. 

Unlike standard text search, which expands terms through ontology mappings and performs spelling corrections, a structured search matches terms **exactly** as written.

To force an exact term search on queries that do not contain explicit boolean operators, set `structured = TRUE`. This bypasses synonym expansion and typo correction:

```{r boolean-forced}
SeqoutSearch("liver cancer", limit = 3)$accession
SeqoutSearch("liver cancer", structured = TRUE, limit = 3)$accession
```

> [!CAUTION]
> Boolean operators are only parsed by full-text search fields. If you combine a boolean query with filters like `assay_l1`, `assay_l2`, or `geo_*`, the search engine will raise an error rather than silently flattening the query.

```{r boolean-clash, error = TRUE}
SeqoutSearch("liver NOT mouse", assay_l1 = "Transcriptomic")
```

## Explore ontology mappings

Standard searches query concepts rather than literal text. For example, a search for `"masld"` automatically matches studies containing `"nonalcoholic fatty liver disease"` because they map to the same concept node in the ontology graph. 

To inspect the synonyms and child terms linked to a search query in the ontology graph, call `Ontology()`:

```{r ontology}
onto <- Ontology("liver")
onto
```

The returned table contains the query term, its synonyms, and its child nodes in the ontology hierarchy.

The `xrefs` column contains ontology identifiers (CURIEs) from standard ontologies (such as UBERON, MeSH, MONDO, and HGNC):

```{r ontology-xrefs}
onto$xrefs[[1]]
```

To find sub-terms, filter for child nodes where `has_children` is `TRUE`:

```{r ontology-children}
kids <- onto[onto$relation == "child" & onto$has_children, "name"]
head(kids, 5)
```

The server caps synonym returns at 500 and child returns at 300. You can check if the results were truncated using the attributes:

```{r ontology-attrs}
attr(onto, "synonym_total")
attr(onto, "children_truncated")
```

The `max_hops` argument (values 1 to 4, default 2) controls the depth of the synonym search walk. If you only need ontology identifiers and want to skip child terms, set `children = FALSE` to reduce query overhead:

```{r ontology-cheap}
Ontology("breast cancer", children = FALSE)$xrefs[[1]]
```

If a term is not present in the ontology graph, the function returns an empty data frame:

```{r ontology-missing}
nrow(Ontology("not a real term at all"))
```

## Sort results and manage limits

By default, the server returns search results ranked by relevance. You can sort results using the `sortby` argument (`"citations"`, `"journal"`, or `"year"`) and set the sort direction using `order` (`"desc"` or `"asc"`):

```{r sortby}
SeqoutSearch("hepatocellular carcinoma", sortby = "citations", limit = 5)
```

### Retrieve total match counts

The server processes search results in pages of 200 rows. The returned object contains the total match count in the `total` attribute and the server execution time in `took_ms`:

```{r all-rows}
result <- SeqoutSearch("hepatocellular carcinoma scRNA")
nrow(result)
attr(result, "took_ms")
```

If you only need a subset of results, set the `limit` argument to reduce the number of pages retrieved:

```{r limit}
nrow(SeqoutSearch("hepatocellular carcinoma scRNA", limit = 50))
```

## Query datasets from search results

After identifying studies using search, pass their accession IDs to `SeqoutGet()` to retrieve their full metadata:

```{r dataset}
top <- SeqoutSearch("hepatocellular carcinoma", sortby = "citations", limit = 1)
d <- SeqoutGet(top$accession[1])

d$meta$title
c(samples = nrow(d$samples), runs = nrow(d$runs))
```

For more details on dataset exploration, see [Metadata](metadata.html).

## Find datasets by publication and author

To find datasets linked to a specific research paper, call `Paper()` and pass either its PubMed ID (`pmid`) or DOI (`doi`):

```{r paper-pmid}
Paper(pmid = "34764296")
```

The `via` column indicates whether the link was declared in the archive deposit or identified by `seqout`.

```{r paper-doi}
Paper(doi = "10.1038/s41467-021-26864-x")
```

If the paper is not linked to any datasets in the database, the function returns an empty data frame:

```{r paper-unknown}
nrow(Paper(pmid = "999999999"))
```

To retrieve all datasets associated with a specific researcher, call `Author()` and pass their full name:

```{r author}
regev <- Author("Aviv Regev")

nrow(regev)
head(regev[, c("accession", "title", "journal", "citation_count")], 5)
```

The name argument must match the spelling in the publication record. Initialized formats (such as `"Regev A"`) are not matched:

```{r author-initials}
nrow(Author("Regev A"))
```

Use `limit` to cap the number of records retrieved (default is 200):

```{r author-limit}
nrow(Author("Aviv Regev", limit = 5))
```

## Construct sample cohorts

To search for individual samples across all studies using harmonized metadata, use `SampleSearch()`. You must supply at least one filter argument.

```{r cohort}
liver <- SampleSearch(
  organism = "Homo sapiens",
  sex = "female",
  tissue = "liver",
  age_min_years = 50
)

head(liver[, c("sample", "study_accession", "tissue", "disease", "assay", "age")])
```

The total number of matching samples is stored in the `total` attribute:

```{r cohort-total}
attr(liver, "total")
```

### Cohort search filters

You can filter samples using the following argument groups:

| Filter Group | Arguments |
| --- | --- |
| Harmonized Attributes | `organism`, `tissue`, `disease`, `cell_type`, `assay`, `sex`, `strain`, `treatment`, and other sample metadata fields. |
| Harmonized Ontology IDs | `disease_ontology_id`, `tissue_ontology_id`, `cell_type_ontology_id`, `assay_ontology_id`, `development_stage_ontology_id`. |
| Value Ranges | `age_min_years`, `age_max_years`, `min_cell_count`, `max_cell_count`, `min_gene_count`, `max_gene_count`. |
| Read-Derived Metrics | `single_cell_only`, `has_viral_reads`, `has_bacterial_reads`, `hpv_type`, `microbe`, `microbe_class`, `microbe_min_breadth`, `microbe_min_kmer_mass`, `microbe_validated_only`. |

Setting `age_min_years = 0` filters for samples that have a recorded age. Samples without age metadata are excluded.

### Query by ontology ID and descendants

To query samples using ontology CURIEs, pass the ID. The search automatically includes samples mapped to descendant subtypes in the ontology graph:

```{r cohort-ontology}
nrow(SampleSearch(disease_ontology_id = "MONDO:0005061", limit = 200))
```

To restrict the search to the exact ontology term and exclude descendant subtypes, set `include_descendants = FALSE`:

```{r cohort-exact}
nrow(SampleSearch(
  disease_ontology_id = "MONDO:0005061",
  include_descendants = FALSE, limit = 200
))
```

### Filter by read-derived microbial screen

You can filter samples based on the presence of specific microbial reads. Specifying the `microbe` argument filters the cohort and attaches the microbial detection counts to the return data frame:

```{r cohort-microbes}
hpv <- SampleSearch(
  tissue = "cervix",
  microbe = "HPV",
  sort = "cell_count",
  order = "desc",
  limit = 10
)

hpv[, c("sample", "study_accession", "cells", "hpv_top_type", "microbe_n_detections")]
```

> [!NOTE]
> Query using the name of the target organism (e.g., `"HPV"`). General terms like `"papillomavirus"` are not matched.

## Execute SQL queries offline

The search and cohort lookup functions require the REST API backend. If you need to perform custom aggregation or query the entire database offline, connect to the Parquet dump backend and write SQL queries directly:

```{r parquet}
con <- SeqoutConnect("parquet", quiet = TRUE)

Query("
  SELECT dominant_scientific_name AS organism, count(*) AS n
  FROM unified_metadata
  WHERE n_samples >= 10
  GROUP BY organism
  ORDER BY n DESC
  LIMIT 10
", con = con)
```

```{r close}
SeqoutClose(con)
```

To download the database dump and configure local connections, see [Downloads](downloads.html).
