---
title: "Search"
description: >
  Search seven archives with one function, filter the studies or the samples,
  and know what a large result set costs.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

There are four ways to search. `SeqoutSearch()` returns studies.
`SampleSearch()` returns samples. `Paper()` and `Author()` search from the
other direction: from a publication to the datasets that it names. All four
read the REST API.

```{r setup}
library(seqout)
```

## A query

`SeqoutSearch()` searches GEO, SRA, ArrayExpress, ENA, GSA, DRA and GEA at the
same time. A query matches the title, the summary and the design of every
study. The server ranks the results.

```{r first-search}
SeqoutSearch("liver cancer scRNA", limit = 5)
```

The `source` column names the archive that holds each result.

## Filters

Give the filters by name. A query is not necessary when you give a filter.

```{r filters-only}
SeqoutSearch(organism = "Mus musculus", assay_l1 = "Transcriptomic", limit = 5)
```

A query and a filter work together.

```{r filters}
SeqoutSearch(
  "liver",
  organism         = "Homo sapiens",
  library_strategy = "RNA-Seq",
  limit            = 5
)
```

## Structured search

A query can be a boolean expression rather than a bag of words. Group terms
with `()`, quote a phrase with `""`, end a term with `*` to match its prefix,
and join them with an uppercase `OR`, `AND` or `NOT`.

```{r boolean}
SeqoutSearch('("aging" OR "aged") (gut OR colon) immun*', limit = 5)
```

Adjacent terms are joined with an implicit `AND`, so `(gut OR colon) immun*`
means "gut or colon, and something beginning with immun".

A structured search takes your terms **exactly**. The ordinary search expands a
query through the ontology graph and corrects a likely typo; a structured one
does neither. That is the point of it: you get what you asked for, and nothing
near it.

The operators are what selects this reading, so you do not have to ask for it.
Only uppercase counts, which leaves ordinary prose alone.

```{r boolean-prose}
nrow(SeqoutSearch("colon or gut", limit = 5))
```

That query has no operators, so it is read as words.

Set `structured = TRUE` to force the exact reading on a query that carries no
operators. The results differ, because the ordinary search expanded the terms
and this one does not.

```{r boolean-forced}
SeqoutSearch("liver cancer", limit = 3)$accession
SeqoutSearch("liver cancer", structured = TRUE, limit = 3)$accession
```

Only the full-text search reads operators. Combined with `assay_l1`,
`assay_l2` or a `geo_*` filter, which no full-text search answers, a boolean
query would come back silently flattened into words, so it is refused instead.

```{r boolean-clash, error = TRUE}
SeqoutSearch("liver NOT mouse", assay_l1 = "Transcriptomic")
```

### The filters

Every filter works with a query and without one, and every filter combines with
every other one.

| Filter | What it selects |
| --- | --- |
| `db`, `source` | one archive |
| `organism` | the organism of the study |
| `library_strategy`, `library_source`, `platform`, `instrument_model` | how the library was made and read |
| `assay_l1`, `assay_l2` | the assay class and the method |
| `country`, `geo_country_code_iso2`, `geo_lat`, `geo_lng`, `geo_radius_km` | where the study comes from |
| `journal` | the journal of the linked paper |
| `multi_platform` | studies that used two or more platforms |
| `date_from`, `date_to` | when the record last changed |

There is one exception. `library_source` cannot be used together with
`assay_l1`, `assay_l2` or a `geo_*` filter, because no search answers both.

```{r clash, error = TRUE}
SeqoutSearch("liver", library_source = "TRANSCRIPTOMIC", assay_l1 = "Transcriptomic")
```

`db` and `source` name the same thing. Use either one to search a single
archive. `db` on its own is the one filter that is not enough: give a query or
one more filter with it.

```{r one-archive}
SeqoutSearch("hepatocellular carcinoma", db = "geo", limit = 5)
```

`assay_l1` and `assay_l2` name the assay. `assay_l1` is the broad class, such
as `"Transcriptomic"`. `assay_l2` is the method, such as `"RNA-seq"`,
`"ATAC-seq"` or `"ChIP-seq"`.

```{r assay}
SeqoutSearch("liver", assay_l2 = "ATAC-seq", limit = 5)
```

### Dates

`date_from` and `date_to` accept dates in `yyyy-mm-dd` format and filter 
datasets by their `updated_at` date.

```{r dates}
SeqoutSearch("liver cancer",
  date_from = "2024-01-01", date_to = "2024-01-31", limit = 5
)
```

A date bound works with any other filter.

```{r dates-combined}
nrow(SeqoutSearch("liver", assay_l1 = "Transcriptomic", date_from = "2025-01-01"))
```

## The vocabulary behind a query

An ordinary search matches concepts, not words: a query for `"masld"` also
finds studies that say `"nonalcoholic fatty liver disease"`, because an
ontology graph joins the two. `Ontology()` shows what the graph joins.

```{r ontology}
onto <- Ontology("liver")
onto
```

The first row is the term. The rows after it are its synonyms, which a search
also matches, and then the terms below it in the hierarchy.

`xrefs` is a list column, because one term can carry an identifier from more
than one ontology. The prefix names the source: UBERON, MeSH, MONDO, HGNC or
Cellosaurus.

```{r ontology-xrefs}
onto$xrefs[[1]]
```

A child with `has_children` set expands further.

```{r ontology-children}
kids <- onto[onto$relation == "child" & onto$has_children, "name"]
head(kids, 5)
```

The server caps the synonyms at 500 and the children at 300. The
`synonym_total` attribute gives the true count, and `children_truncated` says
whether the children hit the cap.

```{r ontology-attrs}
attr(onto, "synonym_total")
attr(onto, "children_truncated")
```

`max_hops` bounds the walk over the synonym links. Set `children = FALSE` to
skip the hierarchy, which is much cheaper when you want the identifiers only.

```{r ontology-cheap}
Ontology("breast cancer", children = FALSE)$xrefs[[1]]
```

A term that the graph does not hold gives an empty tibble, not an error.

```{r ontology-missing}
nrow(Ontology("not a real term at all"))
```

## Sort order

Results come back by relevance. `sortby` takes `"citations"`, `"journal"` and
`"year"`. `order` takes `"desc"`, the default, and `"asc"`.

```{r sortby}
SeqoutSearch("hepatocellular carcinoma", sortby = "citations", limit = 5)
```

## How many rows you get

```{r all-rows}
result <- SeqoutSearch("hepatocellular carcinoma scRNA")
nrow(result)
```

Set `limit` when a sample of the results is sufficient. A limit of 50 costs one
request.

```{r limit}
nrow(SeqoutSearch("hepatocellular carcinoma scRNA", limit = 50))
```

The result carries the server time as an attribute.

```{r took}
attr(result, "took_ms")
```

## From a result to a dataset

A search gives you accessions. `SeqoutGet()` takes one accession and finds
everything below it.

```{r dataset}
top <- SeqoutSearch("hepatocellular carcinoma", sortby = "citations", limit = 1)
d <- SeqoutGet(top$accession[1])

d$meta$title
c(samples = nrow(d$samples), runs = nrow(d$runs))
```

[Metadata](metadata.html) describes the other fields.

## Search by paper

`Paper()` fetched datasets linked to a research paper via its PubMed identifier.

```{r paper-pmid}
Paper(pmid = "34764296")
```

The `via` column says how the link was made, so you can see whether the archive
recorded the paper itself or seqout matched it.

A DOI works the same way. Give one of `pmid` or `doi`, not both.

```{r paper-doi}
Paper(doi = "10.1038/s41467-021-26864-x")
```

A paper that seqout does not hold returns no rows rather than an error, so test
`nrow()` before you use the result.

```{r paper-unknown}
nrow(Paper(pmid = "999999999"))
```

## Search by author

`Author()` takes a name and returns every dataset linked to that author through
its publications.

```{r author}
regev <- Author("Aviv Regev")

nrow(regev)
head(regev[, c("accession", "title", "journal", "citation_count")], 5)
```

The name must match the publication record, which spells authors out in full.
`"Aviv Regev"` works; an initialised form such as `"Regev A"` does not.

```{r author-initials}
nrow(Author("Regev A"))
```

`limit` caps the number of datasets. The default is 200.

```{r author-limit}
nrow(Author("Aviv Regev", limit = 5))
```

A search for studies and a search for authors answer different questions. Use
`SeqoutSearch()` when you know the subject, and `Author()` when you know the
laboratory.

## Sample cohorts

`SampleSearch()` searches samples using their harmonised metadata.

A sample that was never harmonised cannot be found here. At least one filter is
required.

```{r cohort}
liver <- SampleSearch(
  organism = "Homo sapiens",
  sex = "female",
  tissue = "liver",
  age_min_years = 50
)

head(liver[, c("sample", "study_accession", "tissue", "disease", "assay", "age")])
```

The `total` attribute holds the size of the whole cohort, and `filters` holds
the filters that the server applied.

```{r cohort-total}
attr(liver, "total")
```

The filters fall into four groups. The first three read the harmonised data.
The fourth reads the reads.

| Group | Filters |
| --- | --- |
| Harmonised field | `organism`, `tissue`, `disease`, `cell_type`, `assay`, `sex`, `strain`, `treatment`, and the other sample fields |
| Harmonised ontology ID | `disease_ontology_id`, `tissue_ontology_id`, `cell_type_ontology_id`, `assay_ontology_id`, `development_stage_ontology_id` |
| Range | `age_min_years`, `age_max_years`, `min_cell_count`, `max_cell_count`, `min_gene_count`, `max_gene_count` |
| Read-derived | `single_cell_only`, `has_viral_reads`, `has_bacterial_reads`, `hpv_type`, `microbe`, `microbe_class`, `microbe_min_breadth`, `microbe_min_kmer_mass`, `microbe_validated_only` |

An age filter excludes a sample whose age was never recorded. `age_min_years =
0` therefore means "has a recorded age".

### Ontology IDs and their subtypes

Give a harmonised ontology ID as a CURIE. The search expands the term through
the ontology graph, so it also matches the subtypes of that term.

```{r cohort-ontology}
nrow(SampleSearch(disease_ontology_id = "MONDO:0005061", limit = 200))
```

Set `include_descendants = FALSE` for the exact term only.

```{r cohort-exact}
nrow(SampleSearch(
  disease_ontology_id = "MONDO:0005061",
  include_descendants = FALSE, limit = 200
))
```

### What the reads contain

seqout screens the reads themselves. The screen calls the species, the sex and
the assay, and it detects microbial sequence. These calls often disagree with
what the submitter declared, so they are filters of their own.

A `microbe` filter narrows the cohort to the samples that carry a matching
detection, and attaches the detections to each row. One call therefore answers
"cervical single-cell RNA-seq with HPV quantification".

```{r cohort-microbes}
hpv <- SampleSearch(
  tissue = "cervix",
  microbe = "HPV",
  sort = "cell_count",
  order = "desc",
  limit = 10
)

hpv[, c("sample", "study_accession", "cells", "hpv_top_type", "microbe_n_detections")]
```

The screening reference names the organisms HPV16, HPV18 and so on. Search
`"HPV"`, because `"papillomavirus"` matches nothing.

[Metadata](metadata.html) shows the same calls for one study and one sample.

## SQL over the dump

Both search functions need the REST API. For a filter or a count over the whole
index, write SQL against the Parquet dump instead.

```{r parquet}
con <- SeqoutConnect("parquet", quiet = TRUE)

Query("
  SELECT dominant_scientific_name AS organism, count(*) AS n
  FROM unified_metadata
  WHERE n_samples >= 10
  GROUP BY organism
  ORDER BY n DESC
  LIMIT 10
", con = con)
```

```{r close}
SeqoutClose(con)
```
