---
title: "Search Studies and Samples"
description: >
  Learn how to query genomic datasets, apply metadata filters, perform structured 
  boolean searches, resolve ontology terms, and construct sample cohorts.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Four functions search the index:

*   `SeqoutSearch()`: Searches for studies matching text queries or filters.
*   `SampleSearch()`: Searches for individual samples to compile cohorts.
*   `Paper()`: Locates datasets linked to a specific publication.
*   `Author()`: Locates datasets linked to a specific researcher.

All four use the REST API.

```{r setup}
library(seqout)
```

## Search for studies

`SeqoutSearch()` matches study titles, abstracts, and experimental designs across archives.

```{r first-search}
SeqoutSearch("liver cancer scRNA", limit = 5)
```

The `source` column indicates which archive hosts each study.

A misspelled query still searches, but `SearchSuggest()` shows what the server
corrects it to. `SearchFacets()` counts the matches per organism, platform,
country and more without fetching a single row, so you can size a query before
you run it:

```{r suggest-facets}
SearchSuggest("hepatocelular carcinoma")

f <- SearchFacets("hepatocellular carcinoma")
attr(f, "total")
head(f[f$facet == "organism", ], 5)
```

### Apply search filters

Filters narrow results and can be used without query text.

```{r filters-only}
SeqoutSearch(organism = "Mus musculus", assay_l1 = "Transcriptomic", limit = 5)
```

Combine a query with filters:

```{r filters}
SeqoutSearch(
  "liver",
  organism         = "Homo sapiens",
  library_strategy = "RNA-Seq",
  limit            = 5
)
```

### Supported search filters

| Filter | Selection Criteria |
| --- | --- |
| `db` or `source` | One archive (e.g., `geo`, `sra`). |
| `organism` | Scientific name of the target organism. |
| `library_strategy`, `library_source`, `platform`, `instrument_model` | Library preparation and sequencing. |
| `assay_l1`, `assay_l2` | Assay category and specific protocol. |
| `country`, `geo_country_code_iso2`, `geo_lat`, `geo_lng`, `geo_radius_km` | Geographic origin of the study. |
| `journal` | Journal of the linked publication. |
| `multi_platform` | Mapped studies that use more than one sequencing platform. |
| `date_from`, `date_to` | Range of the last modification date (`yyyy-mm-dd`). |

> [!WARNING]
> You cannot combine `library_source` with `assay_l1`, `assay_l2` or any `geo_*` filter, because their search indexes are incompatible.

`db` or `source` needs at least one other search term or filter:

```{r one-archive}
SeqoutSearch("hepatocellular carcinoma", db = "geo", limit = 5)
```

`assay_l1` is a broad category (`"Transcriptomic"`); `assay_l2` is a protocol (`"ATAC-seq"`):

```{r assay}
SeqoutSearch("liver", assay_l2 = "ATAC-seq", limit = 5)
```

### Browse filter vocabularies

Filter values must match the server's spelling. Each vocabulary has a `List*()`
function to look values up.

`ListOrganisms()` returns every organism the index holds, over 300,000 names, so filter it rather than print it:

```{r organisms, eval = FALSE}
organisms <- ListOrganisms()
organisms[grepl("^Mus ", organisms$scientific_name), ]
```

```{r vocabularies}
ListLibraryStrategies()
head(ListInstrumentModels())
ListPlatforms()[, c("platform", "display_name", "total")]
head(ListAssays())
ListCenters(limit = 5)
ListJournals(limit = 5)
```

Country names and ISO 3166-1 alpha-3 codes convert both ways, offline:

```{r countries}
CountryCodeToName(c("IND", "USA"))
CountryNameToCode("India")
```

### Date filtering

```{r dates}
SeqoutSearch("liver cancer",
  date_from = "2024-01-01", date_to = "2024-01-31", limit = 5
)
```

```{r dates-combined}
nrow(SeqoutSearch("liver", assay_l1 = "Transcriptomic", date_from = "2025-01-01"))
```

## Structured boolean searches

Group terms with parentheses `()`, match exact phrases in double quotes `""`, match prefixes with `*`, and join terms with the uppercase operators `AND`, `OR` and `NOT`.

```{r boolean}
SeqoutSearch('("aging" OR "aged") (gut OR colon) immun*', limit = 5)
```

Adjacent terms are joined by an implicit `AND`. A boolean query matches terms exactly, without synonym expansion or spelling correction. `structured = TRUE` does the same for a query without boolean operators:

```{r boolean-forced}
SeqoutSearch("liver cancer", limit = 3)$accession
SeqoutSearch("liver cancer", structured = TRUE, limit = 3)$accession
```

> [!CAUTION]
> Boolean operators are parsed only by full-text search fields. Combining a boolean query with filters like `assay_l1`, `assay_l2`, or `geo_*` raises an error.

```{r boolean-clash, error = TRUE}
SeqoutSearch("liver NOT mouse", assay_l1 = "Transcriptomic")
```

## Control term expansion

The server expands each query term with its synonyms from eight ontologies:
`MONDO`, `MeSH`, `HGNC`, `CHEBI`, `UBERON`, `CL`, `EFO` and `CVCL`. A search for
`"masld"` also finds "nafld".

`expand = FALSE` searches the words as typed, the same request `structured = TRUE`
makes:

```{r expand-off}
nrow(SeqoutSearch("spinal muscular atrophy"))
nrow(SeqoutSearch("spinal muscular atrophy", expand = FALSE))
```

To drop one source's synonyms and keep the rest, name it in `exclude_ontology`:

```{r exclude-ontology}
nrow(SeqoutSearch("spinal muscular atrophy", exclude_ontology = c("MeSH", "CVCL")))
```

A synonym that two ontologies supply stays while either is on, because the graph
holds one node per name carrying the identifiers of every ontology that supplies
it. A name outside the eight raises an error, because the server ignores a name
it does not know.

## Case-sensitive search

Search ignores case, so `"LINE"` also finds every "cell line". Set
`case_sensitive = TRUE` to keep only the studies whose title or summary has
each query word as a whole word, in the case you typed:

```{r case-sensitive}
nrow(SeqoutSearch("LINE", limit = 1000))
nrow(SeqoutSearch("LINE", case_sensitive = TRUE, limit = 1000))
```

The words still follow the query's `OR`, `AND` and `NOT`, and a trailing `*`
still matches a prefix. Spelling correction is off. Filters such as `assay_l1`
and `geo_*` have no case-sensitive match, so combining them with
`case_sensitive` causes an error.

## Explore ontology mappings

`Ontology()` returns a query term, its synonyms and its child terms in the graph:

```{r ontology}
onto <- Ontology("liver")
onto
```

The `xrefs` column holds each row's ontology CURIEs (UBERON, MeSH, MONDO, HGNC and others):

```{r ontology-xrefs}
onto$xrefs[[1]]
```

A child with `has_children` set has sub-terms of its own:

```{r ontology-children}
kids <- onto[onto$relation == "child" & onto$has_children, "name"]
head(kids, 5)
```

The server returns at most 500 synonyms and 300 children. The attributes report the totals and truncation:

```{r ontology-attrs}
attr(onto, "synonym_total")
attr(onto, "children_truncated")
```

`max_hops` (1 to 4, default 2) sets the depth of the synonym walk. `children = FALSE` skips child terms, a cheaper call when you need only identifiers:

```{r ontology-cheap}
Ontology("breast cancer", children = FALSE)$xrefs[[1]]
```

A term missing from the graph returns an empty data frame:

```{r ontology-missing}
nrow(Ontology("not a real term at all"))
```

### Map a metadata column to identifiers

A metadata table holds free text: "T cell", "hepatocyte", "HeLa".
`MapToOntology()` looks each label up in the same graph and puts the
identifiers beside it, in a `<column>_ontology_id` column:

```{r map-to-ontology}
meta <- data.frame(celltype = c("T cell", "hepatocyte", "HeLa"))
MapToOntology(meta, "celltype")
```

`ontology` keeps one source. Pass several column names to map several columns:

```{r map-to-ontology-cl}
MapToOntology(meta, "celltype", ontology = "CL")
```

A label missing from the graph becomes `NA`. Only the label's own identifiers are
read; `use_synonyms = TRUE` lets a label with none take its synonyms' identifiers:

```{r map-to-ontology-synonyms}
MapToOntology(data.frame(celltype = "nbc"), "celltype", use_synonyms = TRUE)
```

Synonym links can join narrower concepts. Each distinct label requires one request.

## Sort results and manage limits

Results are ranked by relevance. `sortby` (`"citations"`, `"journal"` or `"year"`) and `order` (`"desc"` or `"asc"`) change that:

```{r sortby}
SeqoutSearch("hepatocellular carcinoma", sortby = "citations", limit = 5)
```

### Retrieve total match counts

The server returns results in pages of 200 rows. The `total` attribute holds the match count and `took_ms` the server time:

```{r all-rows}
result <- SeqoutSearch("hepatocellular carcinoma scRNA")
nrow(result)
attr(result, "took_ms")
```

`limit` caps the rows, and so the pages fetched:

```{r limit}
nrow(SeqoutSearch("hepatocellular carcinoma scRNA", limit = 50))
```

## Query datasets from search results

`SeqoutGet()` returns a study's full metadata:

```{r dataset}
top <- SeqoutSearch("hepatocellular carcinoma", sortby = "citations", limit = 1)
d <- SeqoutGet(top$accession[1])

d$meta$title
c(samples = nrow(d$samples), runs = nrow(d$runs))
```

See [Metadata](metadata.html) for more.

## Find datasets by publication and author

`Paper()` finds the datasets linked to a paper by PubMed ID (`pmid`) or DOI (`doi`):

```{r paper-pmid}
Paper(pmid = "34764296")
```

The `via` column says whether the archive deposit declared the link or `seqout` found it.

```{r paper-doi}
Paper(doi = "10.1038/s41467-021-26864-x")
```

A paper with no linked datasets returns an empty data frame:

```{r paper-unknown}
nrow(Paper(pmid = "999999999"))
```

`Author()` finds a researcher's datasets by full name:

```{r author}
regev <- Author("Aviv Regev")

nrow(regev)
head(regev[, c("accession", "title", "journal", "citation_count")], 5)
```

The name must match the publication record's spelling. Initials (`"Regev A"`) do not match:

```{r author-initials}
nrow(Author("Regev A"))
```

`limit` caps the records returned (default 200):

```{r author-limit}
nrow(Author("Aviv Regev", limit = 5))
```

## Corpus-wide collections

Each collection has a `*_summary()` (corpus totals), `*_facets()` (filter
values) and `*_projects()` (the studies; `limit = NULL` reads every page):

```{r collections}
# single-cell studies with genetic or chemical perturbation evidence
PerturbationSummary()
pf <- PerturbationFacets()
pf[pf$facet == "perturbation_type", ]
PerturbationProjects(min_confidence = "medium", data_availability = "both", limit = 10)

# spatial transcriptomics studies, by named platform
SpatialSummary()
sf <- SpatialFacets()
head(sf[sf$facet == "platform", ])
SpatialProjects(platform = "Xenium", organism = "Homo sapiens", limit = 10)

# single-cell corpus, any archive
SingleCellSummary()
scf <- SingleCellFacets()
head(scf[scf$facet == "chemistry", ])
SingleCellProjects(chemistry = "10x 3' v3", organism = "Homo sapiens", limit = 10)

# PacBio / Oxford Nanopore studies
LongreadSummary()
lf <- LongreadFacets()
lf[lf$facet == "technology", ]
LongreadProjects(technology = "Oxford Nanopore", limit = 10)

# studies submitted from one country (ISO2 code)
CountrySummary("IN")
cf <- CountryFacets("IN")
head(cf[cf$facet == "organism", ])
CountryProjects("IN", organism = "Homo sapiens", limit = 10)
```

Tissue and disease take a free-text term, resolved against UBERON or MONDO
and expanded to its descendants. `"rare"` and `"nord"` name curated disease
catalogues instead:

```{r ontology-collections}
TissueSummary("liver")
tf <- TissueFacets("liver")
head(tf[tf$facet == "organism", ])
TissueProjects("liver", limit = 10)

DiseaseAliases("marfan", limit = 3) # a colloquial name -> its MONDO ids

DiseaseSummary("rare")
df <- DiseaseFacets("rare")
head(df[df$facet == "category", ])
DiseaseProjects("rare", organism = "Homo sapiens", limit = 10) # or a MONDO term
```

## Construct sample cohorts

`SampleSearch()` finds samples across all studies by harmonized metadata. It needs at least one filter.

```{r cohort}
liver <- SampleSearch(
  organism = "Homo sapiens",
  sex = "female",
  tissue = "liver",
  age_min_years = 50
)

head(liver[, c("sample", "study_accession", "tissue", "disease", "assay", "age")])
```

The `total` attribute holds the match count:

```{r cohort-total}
attr(liver, "total")
```

### Cohort search filters

| Filter Group | Arguments |
| --- | --- |
| Harmonized Attributes | `organism`, `tissue`, `disease`, `cell_type`, `assay`, `sex`, `strain`, `treatment`, and other sample metadata fields. |
| Harmonized Ontology IDs | `disease_ontology_id`, `tissue_ontology_id`, `cell_type_ontology_id`, `assay_ontology_id`, `development_stage_ontology_id`. |
| Value Ranges | `age_min_years`, `age_max_years`, `min_cell_count`, `max_cell_count`, `min_gene_count`, `max_gene_count`. |
| Read-Derived Metrics | `single_cell_only`, `has_viral_reads`, `has_bacterial_reads`, `hpv_type`, `microbe`, `microbe_class`, `microbe_min_breadth`, `microbe_min_kmer_mass`, `microbe_validated_only`. |

`age_min_years = 0` keeps every sample with a recorded age.

### Query by ontology ID and descendants

An ontology ID also matches samples mapped to its descendant subtypes:

```{r cohort-ontology}
nrow(SampleSearch(disease_ontology_id = "MONDO:0005061", limit = 200))
```

`include_descendants = FALSE` matches the exact term only:

```{r cohort-exact}
nrow(SampleSearch(
  disease_ontology_id = "MONDO:0005061",
  include_descendants = FALSE, limit = 200
))
```

### Filter by read-derived microbial screen

`microbe` keeps samples whose reads carry that microbe and attaches the detection counts:

```{r cohort-microbes}
hpv <- SampleSearch(
  tissue = "cervix",
  microbe = "HPV",
  sort = "cell_count",
  order = "desc",
  limit = 10
)

hpv[, c("sample", "study_accession", "cells", "hpv_top_type", "microbe_n_detections")]
```

> [!NOTE]
> Use the organism's name (`"HPV"`). Broader terms like `"papillomavirus"` do not match.

### A worked cohort: HPV16 in single-cell cervical cancer

Add filters one at a time and read `total` to see how each narrows the cohort:

```{r cohort-hpv-funnel}
cervical <- SampleSearch(disease_ontology_id = "MONDO:0002974", limit = 1)
attr(cervical, "total")

sc <- SampleSearch(
  disease_ontology_id = "MONDO:0002974",
  single_cell_only = TRUE,
  limit = 1
)
attr(sc, "total")
```

`MONDO:0002974` is cervical cancer; its subtypes are included by default. `hpv_type` keeps the samples whose reads the Pentimento screen called as that strain:

```{r cohort-hpv16}
hpv16 <- SampleSearch(
  disease_ontology_id = "MONDO:0002974",
  single_cell_only = TRUE,
  hpv_type = "HPV16"
)

hpv16[, c("sample", "study_accession", "disease", "cells", "hpv_top_type")]
```

`hpv_type` selects one strain. `microbe = "HPV"` includes all strains and attaches detections in each sample's `microbes` field:

```{r cohort-hpv-any}
hpv <- SampleSearch(
  disease_ontology_id = "MONDO:0002974",
  single_cell_only = TRUE,
  microbe = "HPV"
)

MicrobeDetections(hpv)
```

Any age filter drops samples without a recorded age, so check coverage first:

```{r cohort-age-coverage}
# age_min_years = 0 means "has a recorded age at all"
with_age <- SampleSearch(
  disease_ontology_id = "MONDO:0002974",
  single_cell_only = TRUE,
  age_min_years = 0,
  limit = 1
)
attr(with_age, "total")
```

`age_days` holds the harmonized age; `age` is the submitter's string, in their own unit.

### The distribution of HPV types across cervical cancer

Drop `single_cell_only`, keep `microbe = "HPV"`, and count the detections. A sample can carry more than one strain, so the counts sum to more than the cohort:

```{r cohort-hpv-distribution}
hpv_all <- SampleSearch(disease_ontology_id = "MONDO:0002974", microbe = "HPV")
c(samples = nrow(hpv_all), studies = length(unique(hpv_all$study_accession)))

all_types <- sort(table(MicrobeDetections(hpv_all)$organism), decreasing = TRUE)
all_types
```

That table counts every k-mer hit, down to a few unitigs against a mostly uncovered genome. `is_viral_evidence` is the screen gate:

```{r cohort-hpv-gated}
gated <- MicrobeDetections(hpv_all, validated_only = TRUE)
gated_types <- sort(table(gated$organism), decreasing = TRUE)
gated_types
```

`breadth_frac` and `kmer_mass` are the gate inputs for custom thresholds.

Both counts on one axis show how much of the tail is trace signal:

```{r cohort-hpv-plot, fig.alt = "Bar chart of HPV type detections across cervical cancer samples, all detections against those passing the viral gate", fig.height = 3.2}
dist <- rbind(
  data.frame(as.data.frame(all_types), set = "any detection"),
  data.frame(as.data.frame(gated_types), set = "passes the gate")
)
names(dist)[1:2] <- c("type", "Freq")

ggplot2::ggplot(dist, ggplot2::aes(stats::reorder(type, Freq), Freq, fill = set)) +
  ggplot2::geom_col(position = "dodge") +
  ggplot2::coord_flip() +
  ggplot2::labs(x = NULL, y = "detections", fill = NULL) +
  ggplot2::theme_minimal()
```

`hpv_top_type` covers samples processed by the Pentimento single-cell pass. `microbes` holds strain detections beyond that coverage.

The denominator includes only screened samples with detections, so these counts cannot estimate HPV prevalence.

`SeqoutGet()` opens the study behind any row and `SeqoutListCounts()` reads its matrices.

## Execute SQL queries offline

Search and cohort functions need the REST API. For custom aggregation, or offline queries over the whole index, connect to the Parquet backend and write SQL:

```{r parquet}
con <- SeqoutConnect("parquet", quiet = TRUE)

Query("
  SELECT dominant_scientific_name AS organism, count(*) AS n
  FROM unified_metadata
  WHERE n_samples >= 10
  GROUP BY organism
  ORDER BY n DESC
  LIMIT 10
", con = con)
```

`Tables()` lists what the dump holds. `Query()` registers the tables a query
names on the fly; `RegisterTables()` does it up front, so dplyr can reach them
through `con$db`:

```{r tables}
head(Tables(con))
con |> RegisterTables("unified_metadata")
dplyr::tbl(con$db, "unified_metadata") |> head(3)
```

Every query re-reads the remote Parquet. `CacheTable()` reads one table once
into the connection as `<table>_local`, so queries against that name skip the
network; `ClearCache()` drops those copies:

```{r cache, eval = FALSE}
CacheTable("geo_series", con = con)
Query("SELECT count(*) FROM geo_series_local", con = con)
ClearCache(con)
```

```{r close}
SeqoutClose(con)
```

To download the dump and set up local connections, see [Downloads](downloads.html).
