---
title: "Search"
description: >
  Search seven archives with one function, filter the studies or the samples,
  and know what a large result set costs.
output:
  rmarkdown::html_document:
    self_contained: true
    mathjax: null
---

```{r, include = FALSE}
seqout::SeqoutDefault(NULL)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

There are two search functions. `SeqoutSearch()` returns studies.
`SampleSearch()` returns samples. Both read the REST API.

```{r setup}
library(seqout)
```

## A query

`SeqoutSearch()` searches GEO, SRA, ArrayExpress, ENA, GSA, DRA and GEA at the
same time. A query matches the title, the summary and the design of every
study. The server ranks the results.

```{r first-search}
SeqoutSearch("liver cancer scRNA", limit = 5)
```

The `source` column names the archive that holds each result.

## Filters

Give the filters by name. A query is not necessary when you give a filter.

```{r filters-only}
SeqoutSearch(organism = "Mus musculus", assay_l1 = "Transcriptomic", limit = 5)
```

A query and a filter work together.

```{r filters}
SeqoutSearch(
  "liver",
  organism         = "Homo sapiens",
  library_strategy = "RNA-Seq",
  limit            = 5
)
```

### The filters

Every filter works with a query and without one, and every filter combines with
every other one.

| Filter | What it selects |
| --- | --- |
| `db`, `source` | one archive |
| `organism` | the organism of the study |
| `library_strategy`, `library_source`, `platform`, `instrument_model` | how the library was made and read |
| `assay_l1`, `assay_l2` | the assay class and the method |
| `country`, `geo_country_code_iso2`, `geo_lat`, `geo_lng`, `geo_radius_km` | where the study comes from |
| `journal` | the journal of the linked paper |
| `multi_platform` | studies that used two or more platforms |
| `date_from`, `date_to` | when the record last changed |

There is one exception. `library_source` cannot be used together with
`assay_l1`, `assay_l2` or a `geo_*` filter, because no search answers both.

```{r clash, error = TRUE}
SeqoutSearch("liver", library_source = "TRANSCRIPTOMIC", assay_l1 = "Transcriptomic")
```

`db` and `source` name the same thing. Use either one to search a single
archive. `db` on its own is the one filter that is not enough: give a query or
one more filter with it.

```{r one-archive}
SeqoutSearch("hepatocellular carcinoma", db = "geo", limit = 5)
```

`assay_l1` and `assay_l2` name the assay. `assay_l1` is the broad class, such
as `"Transcriptomic"`. `assay_l2` is the method, such as `"RNA-seq"`,
`"ATAC-seq"` or `"ChIP-seq"`.

```{r assay}
SeqoutSearch("liver", assay_l2 = "ATAC-seq", limit = 5)
```

The dump lists the values that exist, with a count for each one.

```{r assay-values, eval = FALSE}
Query("
  SELECT dominant_assay_l2 AS assay, count(*) AS n
  FROM unified_metadata
  WHERE assay IS NOT NULL
  GROUP BY assay ORDER BY n DESC
", con = SeqoutConnect("parquet"))
```

### Dates

`date_from` and `date_to` take a day, as `yyyy-mm-dd`. They bound `updated_at`,
which is the day when seqout last saw a change to the record. They are the only
time bounds, and they are the ones the website uses.

```{r dates}
SeqoutSearch("liver cancer",
  date_from = "2024-01-01", date_to = "2024-01-31", limit = 5
)
```

A date bound works with any other filter.

```{r dates-combined}
nrow(SeqoutSearch("liver", assay_l1 = "Transcriptomic", date_from = "2025-01-01"))
```

## Sort order

Results come back by relevance. `sortby` takes `"citations"`, `"journal"` and
`"year"`. `order` takes `"desc"`, the default, and `"asc"`.

```{r sortby}
SeqoutSearch("hepatocellular carcinoma", sortby = "citations", limit = 5)
```

## How many rows you get

A search returns every match. The server sends 200 rows in one page, so a large
result set costs more than one request.

```{r all-rows}
result <- SeqoutSearch("hepatocellular carcinoma scRNA")
nrow(result)
```

Set `limit` when a sample of the results is sufficient. A limit of 50 costs one
request.

```{r limit}
nrow(SeqoutSearch("hepatocellular carcinoma scRNA", limit = 50))
```

The result carries the server time as an attribute.

```{r took}
attr(result, "took_ms")
```

## From a result to a dataset

A search gives you accessions. `SeqoutGet()` takes one accession and finds
everything below it.

```{r dataset}
top <- SeqoutSearch("hepatocellular carcinoma", sortby = "citations", limit = 1)
d <- SeqoutGet(top$accession[1])

d$meta$title
c(samples = nrow(d$samples), runs = nrow(d$runs))
```

[Metadata](metadata.html) describes the other fields.

## Sample cohorts

`SampleSearch()` searches the samples of every study. It reads the ontology
terms that seqout adds to each sample, so a filter can name a tissue, a
disease, an age or a cell count. At least one filter is required.

```{r cohort}
liver <- SampleSearch(
  organism = "Homo sapiens",
  sex = "female",
  tissue = "liver",
  age_min_years = 50
)

head(liver[, c("sample", "study_accession", "tissue", "disease", "assay", "age")])
```

The `total` attribute holds the size of the whole cohort, and `filters` holds
the filters that the server applied.

```{r cohort-total}
attr(liver, "total")
```

The filters fall into four groups.

| Group | Filters |
| --- | --- |
| Text | `organism`, `tissue`, `disease`, `cell_type`, `assay`, `sex`, `strain`, `treatment`, and the other sample fields |
| Ontology term | `disease_ontology_id`, `tissue_ontology_id`, `cell_type_ontology_id`, `assay_ontology_id`, `development_stage_ontology_id` |
| Range | `age_min_years`, `age_max_years`, `min_cell_count`, `max_cell_count`, `min_gene_count`, `max_gene_count` |
| Read-derived | `single_cell_only`, `has_viral_reads`, `has_bacterial_reads`, `hpv_type`, `microbe`, `microbe_class`, `microbe_min_breadth`, `microbe_min_kmer_mass`, `microbe_validated_only` |

An age filter excludes a sample whose age was never recorded. `age_min_years =
0` therefore means "has a recorded age".

### Ontology terms and their subtypes

Give an ontology term as a CURIE. The search expands the term through the
ontology graph, so it also matches the subtypes of that term.

```{r cohort-ontology}
nrow(SampleSearch(disease_ontology_id = "MONDO:0005061", limit = 200))
```

Set `include_descendants = FALSE` for the exact term only.

```{r cohort-exact}
nrow(SampleSearch(
  disease_ontology_id = "MONDO:0005061",
  include_descendants = FALSE, limit = 200
))
```

### What the reads contain

seqout screens the reads themselves. The screen calls the species, the sex and
the assay, and it detects microbial sequence. These calls often disagree with
what the submitter declared, so they are filters of their own.

A `microbe` filter narrows the cohort to the samples that carry a matching
detection, and attaches the detections to each row. One call therefore answers
"cervical single-cell RNA-seq with HPV quantification".

```{r cohort-microbes}
hpv <- SampleSearch(
  tissue = "cervix",
  microbe = "HPV",
  sort = "cell_count",
  order = "desc",
  limit = 10
)

hpv[, c("sample", "study_accession", "cells", "hpv_top_type", "microbe_n_detections")]
```

The screening reference names the organisms HPV16, HPV18 and so on. Search
`"HPV"`, because `"papillomavirus"` matches nothing.

[Metadata](metadata.html) shows the same calls for one study and one sample.

## SQL over the dump

Both search functions need the REST API. For a filter or a count over the whole
index, write SQL against the Parquet dump instead.

```{r parquet}
con <- SeqoutConnect("parquet", quiet = TRUE)

Query("
  SELECT dominant_scientific_name AS organism, count(*) AS n
  FROM unified_metadata
  WHERE n_samples >= 10
  GROUP BY organism
  ORDER BY n DESC
  LIMIT 10
", con = con)
```

```{r close}
SeqoutClose(con)
```
